---
title: "MsQuality – an interoperable open-source package for the calculation of standardized quality metrics of mass spectrometry data"
subtitle: "Supplementary Data"
author:
    - Thomas Naake,^[Genome Biology Unit, European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg, Germany]
    - Johannes Rainer^[Institute for Biomedicine, Eurac Research, Viale Druso 1, 39100 Bolzano, Italy]
    - Wolfgang Huber^[Genome Biology Unit, European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg, Germany]
package: MsQuality

bibliography: document.bib
fontsize: 12pt
output:
  bookdown::pdf_document2:
    includes:
      in_header: "preamble.tex"
    extra_dependencies: ["natbib", "float", "lscape", "subfig", "graphicx"]
    fig_caption: true
    keep_tex: true
    citation_package: natbib
    toc_depth: 3
    number_sections: true
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

\beginsupplement

The Supplementary Information section of this publication demonstrates the 
functionality of the `MsQuality` package by presenting two example analysis 
workflows. The focus is on the data sets of @Cherkaoui2022 and @Amidan2014.

The @Cherkaoui2022 data set is a mass spectrometry metabolomics data set of 
180 cancer cell lines obtained via flow injection analysis 
(TOF, negative ionization mode). The data set comprises a total of 1394 samples.

The @Amidan2014 data set consists of 3400 LC-MS samples of a single QC sample 
(whole cell lysate of *Shewanella oneidensis*). The QC samples were run on 
Exactive, LTQ IonTrap, LTQ Orbitrap, and Velos Orbitrap instruments.

The MsQuality package calculates low-level quality metrics that only require 
minimal information about the mass spectrometry data: retention time, m/z 
values, and associated intensities. The list of quality metrics provided by the
mzQC framework is extensive, also including metrics that depend on more 
high-level information which might not be readily accessible from .raw or 
.mzML files, such as pump pressure mean, or that rely on alignment results, 
like retention time mean shift, signal-to-noise ratio, precursor errors (ppm). 
These metrics are not currently implemented in `MsQuality`. 

The `MsQuality` package is built on the `Spectra` package. Metrics will be 
calculated based on the information stored in a `Spectra` object, and the 
respective `dataOrigin` entries are used to distinguish between the mass 
spectral data of multiple samples.

We would like to note that these metrics only provide an indication of data 
quality, and more advanced analytics, such as those provided by `MatrixQCvis`
[@Naake2022], should be used before removing low-quality samples from 
the analysis. As stated previously [@Bittremieux2017], the type of  QC metrics 
depend on the type of the sample, e.g. if a single peptide or a complex 
lysate of proteins is analzed [@Bereman2015,@Koecher2011,@Paulovich2010]. 
Also, data quality should always be considered in the context of 
the sample type and experimental settings, i.e. quality metrics should always 
be compared with regard to the sample type, experimental setup, instrumentation, 
etc.


In this document, we will (i) create `Spectra` objects from the raw data of the 
two datasets, (ii) calculate the quality metrics on these data sets, 
(iii) visualize some of the metrics, and (iv) assess the performance 
and scalability of the implemented algorithms using the `peakRAM` and 
`microbenchmark` packages.

While the MsQuality package includes an interactive shiny application to 
navigate mass spectral data quality, with plots based on the plotly framework, 
this document will present static plots.
s

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, include=FALSE, echo=FALSE, cache=FALSE}
library("knitr")
knitr::opts_chunk$set(stop_on_error = 1L, fig.pos = "ht")
suppressPackageStartupMessages(library("Spectra"))
suppressPackageStartupMessages(library("MsQuality"))
suppressPackageStartupMessages(library("peakRAM"))
suppressPackageStartupMessages(library("microbenchmark"))
knitr::opts_knit$set(root.dir = "~/Publications/Bioinformatics_MsQuality/")
```

# Preparation of the environment

The following analysis employs functions from several `R` packages, primarily 
the `Spectra` package for representing mass spectrometry spectral data and the 
`MsQuality` package for calculating quality metrics.
Additional packages are necessary for data visualization (`ggplot2`), 
data wrangling (`dplyr`, `readxl``, `stringr`, `tibble`, 
`tidyr`) and performance and scalability analysis
(`peakRAM`, `microbenchmark`).

To proceed with the analysis, please make sure to load these packages before 
starting, for instance using:

```{r load_package, echo = TRUE, eval = TRUE}
library("dplyr")
library("ggplot2")
library("microbenchmark")
library("MsQuality")
library("peakRAM")
library("readxl")
library("Spectra")
library("stringr")
library("tibble")
library("tidyr")
```

\newpage

# Cherkaoui et al. (2022): A functional analysis of 180 cancer cell lines reveals conserved intrinsic metabolic programs

The RAW files were downloaded from the PRIDE database (accession number 
PXD006512, available at www.ebi.ac.uk/pride/archive) via
ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2019/05/PXD006512. They were
then converted into `.mzML` files using MSConvertGUI 
(64-bit, v3.0.22015-aadd392) with the setting `peakPicking` set to 
`vendor msLevel=1-`.

## Prepare the mzML files


In order to read the `mzML` files into `R` and construct a `Spectra` object, 
the `mzML` files need to be modified to conform to a `mzR`-compatible format. 
This is necessary because `mzR` is not able to interpret certain entries and 
references in the original mzML files.

To create `mzR`-compatible `mzML` files, the following steps are taken for each 
`mzML` file in the directory:

- Replace any spaces in the file name with an underscore ("_"),
- Within the `mzML` file, replace the line starting with 
  `<run defaultInstrumentConfigurationRef=` with `<run`,
- Within the `mzML` file, delete the lines between (and including) the lines 
  that start with `<scanWindowList` and `</scanWindowList`
- Remove the original `mzML` file.

These modifications can be automated using the following bash script:

```{bash eval=FALSE, echo=TRUE}
ls *.mzML | while read mzml_i
    do
        new_name="${mzml_i// /_}"
        cp "$mzml_i" "$new_name"
        sed 's/<run defaultInstrumentConfigurationRef=.*/<run/g' "$mzml_i" |
        sed '/^<scanWindowList/,/^<\/scanWindowList/d' > $new_name
        rm "$mzml_i"
    done
```

## Instantiation of the Spectra object

In the subsequent analysis, a `Spectra` object is instantiated. The operations 
were executed within a high-performance computing environment, where the modified 
.mzML files were stored in the directory `Cherkaoui2022`.

```{r echo = FALSE}
setwd("~/GitHub/MsQuality_manuscript/")
```

```{r create_spectra_Cherkaoui, eval = FALSE, echo = TRUE, cache=TRUE}
## read the file with protein intensities
.path <- "/scratch/naake/Cherkaoui2022/"
fls <- dir(.path, full.names = TRUE, pattern = "mzML")

## create the Spectra object
sps <- Spectra(fls, backend = MsBackendMzR())
```

```{r sps_saveRDS_Cherkaoui2022, echo = FALSE, eval = FALSE}
## save the Spectra object as RDS file
saveRDS(sps, file = "Cherkaoui2022/Cherkaoui2022_sps.RDS")
```

## Calculate the metrics via `MsQuality`

`MsQuality` utilizes `Spectra` objects that store the spectral data. In 
this particular case, where the spectral data was obtained via flow injection 
analysis, metrics that incorporate retention time information are not 
relevant and the analysis will only focus on the three metrics

 - `numberSpectra`, **Number of MS1 spectra** (QC:4000059), “The number of MS1 events in the run.” [PSI:QC];
 - `areaUnderTic`, **Area under TIC** (QC:4000077), “The area under the total ion chromatogram.” [PSI:QC];
 - `mzAcquisitionRange`, **m/z acquisition range** (QC:4000138), “Upper and lower limit of m/z values at which spectra are recorded.” [PSI:QC].


The calculation of the metrics is achieved using the function 
`calculateMetricsFromSpectra`, which takes as input the `Spectra` object, 
`sps`, and the above-defined metrics. Additionally, optional parameters 
can be passed to `calculateMetricsFromSpectra` to further specify the 
calculation, such as the `msLevel` if multiple mass spectra levels are present 
in the `Spectra` object. In the current context, where only MS1 level spectra 
are stored in the `Spectra` object, specifying the `msLevel` is not necessary.

```{r calculateMetrics_Cherkaoui2022, eval = FALSE}
.metrics <- c("numberSpectra", "areaUnderTic", "mzAcquisitionRange")

.metrics_sps <- calculateMetricsFromSpectra(spectra = sps, 
    metrics = .metrics)
```

```{r .metrics_sps_saveRDS_Cherkaoui2022, eval = FALSE}
saveRDS(.metrics_sps, file = "Cherkaoui2022/Cherkaoui2022_metrics_sps.RDS")
```

```{r .metrics_sps_readRDS_saveRDS_Cherkaoui2022, eval = TRUE, echo = FALSE}
.metrics_sps <- readRDS("Cherkaoui2022/Cherkaoui2022_metrics_sps.RDS")
```

## Visualization


In the analysis of the @Cherkaoui2022 study, the three quality metrics were
visualized using the` ggplot2` package. The XLSX file metabolomics_180CCL.xlsx,
downloaded from
https://www.research-collection.ethz.ch/handle/20.500.11850/511784, was used 
to extract information on the samples that were included in the subsequent 
analysis of the @Cherkaoui2022 study. This information was added to the 
`.metrics_sps` object, specifically, whether the sample was 
analyzed (`"yes"`) or excluded (`"no"`. 
A plot was then created to compare the differences in quality metrics 
between the analyzed and excluded samples.

```{r}
## reshape the metrics into long format
.metrics_sps <- .metrics_sps |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble() |>
    pivot_longer(cols = 2:(ncol(.metrics_sps) + 1))

## for visualization purposes truncate the dataOrigin, the rowname entry
## will contain 
.metrics_sps[["rowname"]] <- .metrics_sps[["rowname"]] |>
    strsplit(split = "Cherkaoui2022/") |>
    lapply(FUN = function(names_i) names_i[[2]]) |> 
    unlist() |>
    str_remove(pattern = ".mzML")

## add information if the samples was analyzed (information stored in 
## .samples)
.samples <- read_excel("Cherkaoui2022/PrimaryAnalysis/metabolomics_180CCL.xlsx",
    sheet = "injections")
.metrics_sps <- .metrics_sps |> 
    mutate(dsCode = sapply(
        strsplit(.metrics_sps$rowname, split = "_Batch"), "[", 1)) |>
    mutate(analyzed = ifelse(dsCode %in% .samples[["dsCode"]], "yes", "no"))
```

```{r, fig.cap = "Quality metrics for data set of Cherkaoui et al. (2002) stratified by information if the sample was analyzed or excluded. Displayed are the area under the TIC (areaUnderTic) and the minimum and maximum values of the m/z acquisition range (mzAcquisitionRange.min, mzAcquisitionRange.max)."}
subset(.metrics_sps, 
    name %in% c("areaUnderTic", "mzAcquisitionRange.min", "mzAcquisitionRange.max")) |>
    ggplot() +
        geom_violin(aes(x = analyzed, y = value, col = analyzed)) +
        geom_jitter(aes(x = analyzed, y = value, col = analyzed)) +
        facet_wrap(~ name, scales = "free") +
        theme_classic()
```

The plot demonstrates that the excluded samples have lower total ion current 
(TIC) values, which was already noted in the original publication and was the 
reason for their exclusion from subsequent analysis steps. The plot serves as 
a visual confirmation of this statement and aids in understanding the data 
quality of the samples.

## Calculate Peak RAM Used

An important aspect, especially when dealing with large amount of data,
is scalability and performance when computing the quality metric. In this 
section, the total and peakRAM is monitored via `peakRAM`. 
This package allows for a detailed analysis of the memory usage during the 
execution of code, providing insights on potential performance bottlenecks and 
opportunities for optimization. By monitoring memory usage, it is possible 
to determine the scalability of the code and ensure that the performance of
the analysis remains acceptable as the data size increases.

```{r df_ram_Cherkaoui2022, eval = FALSE, echo = TRUE}
.path <- "/scratch/naake/Cherkaoui2022"
fls <- dir(.path, full.names = TRUE, pattern = "mzML")
.metrics <- c("numberSpectra", "areaUnderTic", "mzAcquisitionRange")
df_ram <- peakRAM(
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 1))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 2))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 4))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 8))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 16))
	}
)
```


```{r df_ram_saveRDS_Cherkaoui2022, eval = FALSE, echo = FALSE}
## save the data.frame as RDS object
saveRDS(df_ram, file = "Cherkaoui2022/Cherkaoui2022_df_ram.RDS")
```

```{r df_ram_readRDS_Cherkaoui2022, eval = TRUE, echo = FALSE}
df_ram <- readRDS("Cherkaoui2022/Cherkaoui2022_df_ram.RDS")
```

```{r df_ram_plot_Cherkaoui2022, fig.cap = "Total and peak RAM for the calculation of quality metrics of the data set of Cherkaoui et al. (2022) under parallelization (1, 2, 4, 8, and 16 workers)."}
df_ram[["Function_Call"]] <- c(1, 2, 4, 8, 16)

df_ram <- df_ram |> 
    pivot_longer(cols = 2:ncol(df_ram), names_to = "parameter") |>
    mutate(parameter = case_when(
        parameter == "Elapsed_Time_sec" ~ "elapsed time (sec)",
        parameter == "Total_RAM_Used_MiB" ~ "used total RAM",
        parameter == "Peak_RAM_Used_MiB" ~ "used peak RAM"))

## visualize the elapsed time per amount of used workers
df_ram |>
    filter(parameter != "elapsed time (sec)") |>
    ggplot() +
        geom_point(aes(x = Function_Call, y = value), size = 3) +
        xlab("number of workers") + ylab("MiB") +
        scale_x_continuous(breaks = df_ram[["Function_Call"]]) +
        facet_wrap(~ parameter, scales = "free_y") +
        theme_bw() +
        theme(axis.title = element_text(size = 16), 
            axis.text = element_text(size = 14),
            panel.grid = element_blank())
```

We observe here that overhead is controlled and the total and peak RAM does not 
scale with the number of cores. In general the total and peak RAM was low and 
managable by a personal computer.

In another application, we measured the time it takes to evaluate the 
calculation of quality metrics under parallelizing the tasks on
1, 2, 4, 8, and 16 workers using the `microbenchmark` package. This package 
allows for precise measurement of the execution time of `R` expressions by 
repeating the evaluation multiple times and providing detailed summary 
statistics of the execution times. The parallelization process can help in 
the management of bigger data sets, and to save valuable time in data analysis.

```{r df_mb_Cherkaoui2022, echo = TRUE, eval = FALSE}
library("microbenchmark")
.path <- "/scratch/naake/Cherkaoui2022"
fls <- dir(.path, full.names = TRUE, pattern = "mzML")
fls <- fls[1:1000]
.metrics <- c("numberSpectra", "areaUnderTic", "mzAcquisitionRange")
df_mb <- microbenchmark(
    workers_1 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 1)),
    workers_2 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 2)),
    workers_4 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 4)),
    workers_8 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 8)),
    workers_16 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 16)), 
	times = 110L, control = list(warmup = 10), check = "equal"
)
```

```{r df_mb_saveRDS_Cherkaoui2022, eval = FALSE, echo = FALSE}
## save the data.frame
saveRDS(df_mb, file = "Cherkaoui2022/Cherkaoui2022_df_mb.RDS")
```

```{r df_mb_readRDS_Cherkaoui2022, eval = TRUE, echo = FALSE}
df_mb <- readRDS(file = "Cherkaoui2022/Cherkaoui2022_df_mb.RDS")
```

```{r df_mb_plot_Cherkaoui2022, echo = TRUE, eval = TRUE, fig.cap = "Execution time for the calculation of quality metrics of the data set of Cherkaoui et al. (2022) under parallelization (1, 2, 4, 8, and 16 workers)."}
## convert from nano seconds to seconds
df_mb[["time"]] <- df_mb[["time"]] / 10e9
df_mb <- df_mb |>
    as.data.frame() |>
    mutate(expr = case_when(expr == "workers_1" ~ 1,
        expr == "workers_2" ~ 2, expr == "workers_4" ~ 4,
        expr == "workers_8" ~ 8, expr == "workers_16" ~ 16))

## visualize the elapsed time per amount of used workers
ggplot(df_mb) +
    geom_jitter(aes(x = expr, y = time, col = as.factor(expr)), alpha = 0.2) +
    geom_violin(aes(x = expr, y = time, group = expr, fill = NA), fill = alpha("white", 0)) +
    xlab("number of workers") + ylab("elapsed time (sec)") +
    scale_x_continuous(breaks = df_mb[["expr"]]) +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    theme_classic() +
    theme(axis.title = element_text(size = 16), 
            axis.text = element_text(size = 14))
```

By parallelizing the calculation of the 
quality metrics across multiple workers, it is possible to significantly 
reduce the execution time, and the microbenchmark package was used to 
accurately measure the performance improvements achieved by parallelization. 

# Amidan et al. (2014): Signatures for mass spectrometry data quality


The RAW files were downloaded from 

- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000320
  (`1_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000321
  (`2_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000322
  (`3_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000323
  (`4_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000324
  (`5_of_5`).

Subsequently, the RAW files were converted into `.mzML` files using 
MSConvertGUI (64-bit, v3.0.22015-aadd392) with setting
`peakPicking` to `vendor msLevel=1-`.


## Instantiation of the Spectra object

In the subsequent analysis, a `Spectra` object is instantiated. The operations 
were executed within a high-performance computing environment, where the modified 
.mzML files were stored in the directory `Amidan2014`.

```{r create_spectra_Amidan2014, echo = TRUE, eval = FALSE}
## read the file with protein intensities
.path <- "/scratch/naake/Amidan2014"
fls <- dir(.path, full.names = TRUE, recursive = TRUE, pattern = "mzML")

## create the Spectra object
sps <- Spectra(fls, backend = MsBackendMzR())
```

```{r sps_saveRDS_Amidan2014, echo = FALSE, eval = FALSE}
## save the Spectra object as RDS file
saveRDS(sps, file = "Amidan2014/Amidan2014_sps.RDS")
```

## Calculate the metrics via `MsQuality`


`MsQuality` utilizes `Spectra` objects that store the spectral data. Here,
retention time information was available from the mzML files and a higher
number of metrics could be calculated:

- *rtDuration*, **RT duration** (QC:4000053), "The retention time duration of 
  the MS run in seconds, similar to the highest scan time minus the lowest 
  scan time." [PSI:QC];
- *rtOverTicQuantile*, **RT over TIC quantile** (QC:4000054), "The interval 
  when the respective quantile of the TIC accumulates divided by retention time 
  duration. The number of quantiles observed is given by the size of the tuple." 
  [PSI:QC];
- *rtOverMsQuarters*, **MS1 quantiles RT fraction** (QC:4000055), 
  "The interval used for acquisition of the first, second, third, and fourth 
  quarter of all MS1 events divided by RT-Duration." [PSI:QC];
- *rtOverMsQuarters*, **MS2 quantiles RT fraction** (QC:4000055), 
  "The interval used for acquisition of the first, second, third, and fourth 
  quarter of all MS2 events divided by RT-Duration." [PSI:QC];
- *ticQuartileToQuartileLogRatio*, **MS1 TIC-change quartile ratios** 
  (MS:4000057), "The log ratios of successive TIC-change quartiles. The 
  TIC changes are the list of MS1 total ion current (TIC) value changes 
  from one to the next scan, produced when each MS1 TIC is subtracted 
  from the preceding MS1 TIC. The metric's value triplet represents the 
  log ratio of the TIC-change Q2 to Q1, Q3 to Q2, TIC-change-max to Q3" 
  [PSI:MS], `mode = "TIC_change"`;
- *ticQuartileToQuartileLogRatio*, **MS1 TIC quartile ratios** (MS:4000058),
  The log ratios of successive TIC quartiles. The metric's value triplet 
  represents the log ratios of TIC-Q2 to TIC-Q1, TIC-Q3 to TIC-Q2, TIC-max to 
  TIC-Q3." [PSI:MS], `mode = "TIC";
- *numberSpectra*, **Number of MS1 spectra** (QC:4000059), 
  "The number of MS1 events in the run." [PSI:QC]; 
- *numberSpectra*, **Number of MS2 spectra** (QC:4000060), 
  "The number of MS2 events in the run." [PSI:QC];
- *medianPrecursorMz*, **Precursor median m/z for IDs** (QC:4000065), 
  "Median m/z value for all identified peptides (unique ions) after FDR." 
  [PSI:QC];
- *rtIqr*, **Interquartile RT period for peptide identifications** (QC:4000072), 
  "The interquartile retention time period, in seconds, for all peptide 
  identifications over the complete run." [PSI:QC];
- *rtIqrRate*, **Peptide identification rate of the interquartile RT period** 
  (QC:4000073), "The identification rate of peptides for the interquartile 
  retention time period, in peptides per second." [PSI:QC];
- *areaUnderTic*, **Area under TIC** (QC:4000077), "The area under the total ion
  chromatogram." [PSI:QC];
- *areaUnderTicRtQuantiles*, **Area under TIC RT quantiles** (QC:4000078), "The 
  area under the total ion chromatogram of the retention time quantiles. Number 
  of quantiles are given by the n-tuple." [PSI:QC];
- *medianTicRtIqr*, **Median of TIC values in the RT range in which the middle half of peptides are identified** 
  (QC:4000130), "Median of TIC values in the RT range in which half of peptides 
  are identified (RT values of Q1 to Q3 of identifications)" [PSI:QC];
- *medianTicOfRtRange*, **Median of TIC values in the shortest RT range in which half of the peptides are identified** 
  (QC:4000132), "Median of TIC values in the shortest RT range in which half of 
  the peptides are identified"  [PSI:QC] 
- *mzAcquisitionRange*, **m/z acquisition range** (QC:4000138), "Upper and 
  lower limit of m/z values at which spectra are recorded." [PSI:QC];
- *rtAcquisitionRange*, **Retention time acquisition range** (QC:4000139), 
  "Upper and lower limit of time at which spectra are recorded." [PSI:QC];
- *precursorIntensityRange*, **Precursor intensity range** (QC:4000144), 
  "Minimum and maximum precursor intensity recorded." [PSI:QC];
- *precursorIntensityQuartiles*, **Precursor intensity distribution Q1, Q2, Q3**
  (QC:4000167), "From the distribution of precursor intensities, the quartiles 
Q1, Q2, Q3" [PSI:QC];
- *precursorIntensityMean*, **Precursor intensity distribution mean** 
  (QC:4000168), "From the distribution of precursor intensities, the mean." 
  [PSI:QC];
- *precursorIntensitySd*, **Precursor intensity distribution sigma** 
  (QC:4000169), "From the distribution of precursor intensities, the sigma 
  value." [PSI:QC];
- *msSignal10xChange*, **MS1 signal jump (10x) count** (QC:4000172), "The 
  count of MS1 signal jump (spectra sum) by a factor of ten or more (10x) 
  between two subsequent scans" [PSI:QC];
- *RatioCharge1over2*, **Charged peptides ratio 1+ over 2+** (QC:4000174), 
  "Ratio of 1+ peptide count over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge1over2*, **Charged spectra ratio 1+ over 2+** (QC:4000179), 
  "Ratio of 1+ spectra count over 2+ spectra count in all MS2" [PSI:QC];
- *RatioCharge3over2*, **Charged peptides ratio 3+ over 2+** (QC:4000175), 
  "Ratio of 3+ peptide count over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge3over2*, **Charged spectra ratio 3+ over 2+** (QC:4000180), 
  "Ratio of 3+ peptide count over 2+ peptide count in all MS2" [PSI:QC];
- *RatioCharge4over2*, **Charged peptides ratio 4+ over 2+** (QC:4000176), 
  "Ratio of 4+ peptide count  over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge4over2*, **Charged spectra ratio 4+ over 2+** (QC:4000181), 
  "Ratio of 4+ peptide count over 2+ peptide count in all MS2" [PSI:QC];
- *meanCharge*, **Mean charge in identified spectra** (QC:4000177), 
  "Mean charge in identified spectra" [PSI:QC];
- *meanCharge*, **Mean precursor charge in all MS2** (QC:4000182), 
  "Mean precursor charge in all MS2" [PSI:QC];
- *medianCharge*, **Median charge in identified spectra** (QC:4000178), 
  "Median charge in identified spectra" [PSI:QC];
- *medianCharge*, **Median precursor charge in all MS2** (QC:4000183), 
  "Median precursor charge in all MS2" [PSI:QC].

Again, the calculation of the metrics is achieved using the function
`calculateMetricsFromSpectra`, which takes as input the `Spectra` object, 
`sps`, and the above-defined metrics. We calculate some of the metrics
based on their respective `msLevel` and decide here to calculate the metrics
separately for `msLevel 1` and `msLevel 2`.

Overall, this function provides a flexible and 
efficient way to analyze large amounts of mass spectrometry data and obtain 
insights on the quality of the data.

```{r, eval = FALSE}
.metrics <- c("rtDuration", "rtOverTicQuantile", "rtOverMsQuarters", 
    "ticQuartileToQuartileLogRatio", "numberSpectra", 
    "medianPrecursorMz", "rtIqr", "rtIqrRate", "areaUnderTic", 
    "areaUnderTicRtQuantiles", "medianTicRtIqr", "medianTicOfRtRange",
    "mzAcquisitionRange", "rtAcquisitionRange", "precursorIntensityRange", 
    "precursorIntensityQuartiles", "precursorIntensityMean", 
    "precursorIntensitySd", "msSignal10xChange", "ratioCharge1over2", 
    "ratioCharge3over2", "ratioCharge4over2", "meanCharge", 
    "medianCharge")

.metrics_sps_msLevel1 <- MsQuality::calculateMetricsFromSpectra(spectra = sps, 
    metrics = .metrics, msLevel = 1L)
.metrics_sps_msLevel1 <- MsQuality::calculateMetricsFromSpectra(spectra = sps, 
    metrics = .metrics, msLevel = 2L)
```


```{r .metrics_sps_saveRDS_Amidan2014, echo = FALSE, eval = FALSE}
saveRDS(.metrics_sps_msLevel1, 
    file = "Amidan2014/Amidan2014_metrics_sps_msLevel1.RDS")
saveRDS(.metrics_sps_msLevel2, 
    file = "Amidan2014/Amidan2014_metrics_sps_msLevel2.RDS")
```

```{r .metrics_sps_readRDS_Amidan2014, echo = FALSE, eval = TRUE}
.metrics_sps_msLevel1 <- readRDS("Amidan2014/Amidan2014_metrics_sps_msLevel1.RDS")
.metrics_sps_msLevel2 <- readRDS("Amidan2014/Amidan2014_metrics_sps_msLevel2.RDS")
```

## Visualization

In the analysis of the @Amidan2014 study, the quality metrics were
visualized using the` ggplot2` package. The XLS files pr401143e_si_002.xls
and pr401143e_si_003.xls (provided as Supplemental Material of the original
publication) was used to extract information on the sample quality.
This information was added to the `.metrics_sps` object.
A plot was then created to compare the differences in quality metrics 
between the low- and high-quality samples samples. 

```{r join_metrics_and_metadata}
## reshape the metrics into long format
.metrics_sps_msLevel1 <- .metrics_sps_msLevel1 |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble()
.metrics_sps_msLevel2 <- .metrics_sps_msLevel2 |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble()

## for visualization purposes truncate the dataOrigin, the rowname entry
## will contain 
.metrics_sps_msLevel1[["rowname"]] <- .metrics_sps_msLevel1[["rowname"]] |>
    strsplit(split = "Amidan2014/[1-5]_of_5/", fixed = FALSE) |>
    lapply(FUN = function(names_i) names_i[[2]]) |> 
    unlist() |>
    str_remove(pattern = ".mzML")
.metrics_sps_msLevel2[["rowname"]] <- .metrics_sps_msLevel2[["rowname"]] |>
    strsplit(split = "Amidan2014/[1-5]_of_5/", fixed = FALSE) |>
    lapply(FUN = function(names_i) names_i[[2]]) |> 
    unlist() |>
    str_remove(pattern = ".mzML")

## add information if the samples was analyzed (information stored in 
## .samples)
##add here information on the classification of samples
.samples_training <- read_excel("Amidan2014/pr401143e_si_002.xls",
    sheet = "FileS1_TrainingDataset")
.samples_validation <- read_excel("Amidan2014/pr401143e_si_003.xls",
    sheet = "FileS2_ValidationDataset")
.cols <- intersect(colnames(.samples_training), colnames(.samples_validation))
.samples <- rbind(.samples_training[, .cols], .samples_validation[, .cols])

.metrics_sps_msLevel1 <- left_join(.metrics_sps_msLevel1, .samples, 
    by = c("rowname" = "Dataset"), copy = TRUE)
.metrics_sps_msLevel2 <- left_join(.metrics_sps_msLevel2, .samples, 
    by = c("rowname" = "Dataset"), copy = TRUE)
```

```{r areaUnderTIC, fig.cap = "Quality metrics by MsQuality: Area under TIC. The metrics are calculated from MS1 and MS2 spectra."}
## area under TIC
gg_tic_msLevel1 <- .metrics_sps_msLevel1 |>
    select(rowname, Curated_Quality, rtDuration, areaUnderTic) |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = areaUnderTic) |>
    ggplot() +
        stat_density_2d(aes(x = rtDuration, y = value, color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        xlab("RT duration (s, in 10e3)") + ylab("area under TIC (MS1)") +
        theme_classic() +
        theme(legend.position = "none")

gg_tic_msLevel2 <- .metrics_sps_msLevel2 |>
    select(rowname, Curated_Quality, rtDuration, areaUnderTic) |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = areaUnderTic) |>
    ggplot() +
        stat_density_2d(aes(x = rtDuration, y = value, color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        xlab("RT duration (s, in 10e3)") + ylab("area under TIC (MS2)") +
        theme_classic() +
        theme(legend.position = "none")

ggpubr::ggarrange(gg_tic_msLevel1, gg_tic_msLevel2, ncol = 2, nrow = 1, labels = "AUTO")
```

```{r, fig.cap = "Quality metrics by MsQuality: Area under TIC RT quantiles. The metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities stratified for high-quality and low-quality samples as classified in Amidan et al. (2014). A: 25% quantile for MS1 spectra. B: 50% quantile for MS1 spectra. C: 75% quantile for MS1 spectra. D: 100% quantile for MS1 spectra. E: 25% quantile for MS2 spectra. F: 50% quantile for MS2 spectra. G: 75% quantile for MS2 spectra. H: 100% quantile for MS2 spectra."}
## areaUnderTicRtQuantiles
## areaUnderTicRtQuantiles 25%
gg_tic25_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.25%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.25%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("25% (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic25_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.25%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.25%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("25% (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

## areaUnderTicRtQuantiles 50%
gg_tic50_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.50%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.50%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("50% (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic50_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.50%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.50%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("50% (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

## areaUnderTicRtQuantiles 75%
gg_tic75_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.75%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.75%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("75% (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic75_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.75%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.75%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("75% (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

## areaUnderTicRtQuantiles 100%
gg_tic100_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.100%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.100%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("100% (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic100_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "areaUnderTicRtQuantiles.100%") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "areaUnderTicRtQuantiles.100%") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("100%  (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

p <- ggpubr::ggarrange(gg_tic25_msLevel1, gg_tic50_msLevel1, 
    gg_tic75_msLevel1, gg_tic100_msLevel1, 
    gg_tic25_msLevel2, gg_tic50_msLevel2, 
    gg_tic75_msLevel2, gg_tic100_msLevel2, 
    ncol = 4, nrow = 2, labels = "AUTO")

ggpubr::annotate_figure(p, 
    top = ggpubr::text_grob("area under TIC RT quantile"),
    bottom = "RT duration (s, in 10e3)")
```

```{r ticQuartileToQuartileLogRatio, fig.cap = "Quality metrics by MsQuality: TIC quantile to quantile log ratio. The metrics are calculated from MS1 and MS2 spectra."}
## ticQuartileToQuartileLogRatio
## ticQuartileToQuartileLogRatio Q2/Q1%
gg_tic_q2_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q2/Q1") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q2/Q1") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q2/Q1 (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic_q2_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q2/Q1") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q2/Q1") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q2/Q1 (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

## ticQuartileToQuartileLogRatio.Q3/Q2
gg_tic_q3_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q3/Q2") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q3/Q2") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q3/Q2 (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic_q3_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q3/Q2") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q3/Q2") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q3/Q2 (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

## ticQuartileToQuartileLogRatio.Q4/Q3
gg_tic_q4_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q4/Q3") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q4/Q3") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q4/Q3 (MS1)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

gg_tic_q4_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!duplicated(rowname)) |>
    select(rowname, Curated_Quality, rtDuration, "ticQuartileToQuartileLogRatio.Q4/Q3") |>
    filter(Curated_Quality %in% c("good", "poor")) |>
    mutate(rtDuration = rtDuration / 1000) |>
    pivot_longer(cols = "ticQuartileToQuartileLogRatio.Q4/Q3") |>
    ggplot(aes(x = rtDuration, y = value)) +
        stat_density_2d(aes(color = Curated_Quality)) +
        facet_wrap(~ Curated_Quality, scales = "fixed") +
        ylab("Q4/Q3 (MS2)") +
        theme_classic() +
        theme(legend.position = "none", axis.title.x = element_blank())

p <- ggpubr::ggarrange(gg_tic_q2_msLevel1, gg_tic_q3_msLevel1, gg_tic_q4_msLevel1, 
    gg_tic_q2_msLevel2, gg_tic_q3_msLevel2, gg_tic_q4_msLevel2, 
    ncol = 3, nrow = 2, labels = "AUTO")

ggpubr::annotate_figure(p, 
    top = ggpubr::text_grob("TIC quantile to quantile log ratio"),
    bottom = "RT duration (s, in 10e3)")
```

The plot demonstrates that the low-quality samples (poor) have lower total ion 
current (TIC) values compared to high-quality samples (good). The plots serve as
a visual check to differences in TIC values and aids in understanding the data
quality of the samples.

### Comparison to QuaMeter metrics

The XLS files also contain pre-calculated `QuaMeter` metrics [@Ma2012] for each of 
the samples. In the following, we will compare the QuaMeter-metrics to the
`msQuality` metrics to check if the `MsQuality` shows concordant results
compared to the `QuaMeter` metrics.

```{r rtDuration, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: RT duration. The metrics by MsQuality are calculated from MS1 and MS2 spectra."}
## rtDuration (MsQuality) vs. RT_Duration (QuaMeter)
rt_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = rtDuration, y = RT_Duration)) +
        geom_point(alpha = 0.2) +
        xlab("MsQuality (MS1)") + ylab("QuaMeter") +
        theme_classic() +
        xlim(0, 40000) + ylim(0, 40000) +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

rt_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = rtDuration, y = RT_Duration)) +
        geom_point(alpha = 0.2) +
        xlab("MsQuality (MS2)") + ylab("QuaMeter") +
        theme_classic() +
        xlim(0, 40000) + ylim(0, 40000) +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

p <- ggpubr::ggarrange(rt_msLevel1, rt_msLevel2, ncol = 2, nrow = 1, labels = "AUTO")
ggpubr::annotate_figure(p, top = ggpubr::text_grob("RT duration (s)"))
```

```{r numberSpectra, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: Number of spectra. The metrics are calculated from MS1 and MS2 spectra."}
## 
##  numberSpectra (MsQuality) vs. MS1_Count (QuaMeter)
.metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = numberSpectra, y = MS1_Count)) +
        geom_point(alpha = 0.2) +
        xlab("numberSpectra (MS1, MsQuality)") + ylab("MS1 count (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

##  numberSpectra (MsQuality) vs. MS2_Count (QuaMeter)
.metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = numberSpectra, y = MS2_Count)) +
        geom_point(alpha = 0.2) +
        xlab("numberSpectra (MS2, MsQuality)") + ylab("MS2 count (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))
```

```{r rtOverTicQuantile, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: RT over TIC quantile. The metrics are calculated from MS1 and MS2 spectra."}
## 
##  rtOverTicQuantile (MsQuality) vs. RT_TIC_Q1 (QuaMeter)
gg_rt_tic_q1_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q1 = RT_TIC_Q1) |>
    ggplot(aes(x = `rtOverTicQuantile.25%`, y = RT_TIC_Q1)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 25% (MS1, MsQuality)") + ylab("RT_TIC_Q1 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_tic_q1_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q1 = RT_TIC_Q1) |>
    ggplot(aes(x = `rtOverTicQuantile.25%`, y = RT_TIC_Q1)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 25% (MS2, MsQuality)") + ylab("RT_TIC_Q1 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

##  rtOverTicQuantile (MsQuality) vs. RT_TIC_Q2 (QuaMeter)
gg_rt_tic_q2_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    ggplot(aes(x = `rtOverTicQuantile.50%`, y = RT_TIC_Q2)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 50% (MS1, MsQuality)") + ylab("RT_TIC_Q2 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_tic_q2_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    ggplot(aes(x = `rtOverTicQuantile.50%`, y = RT_TIC_Q2)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 50% (MS2, MsQuality)") + ylab("RT_TIC_Q2 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

##  rtOverTicQuantile (MsQuality) vs. RT_TIC_Q3 (QuaMeter)
gg_rt_tic_q3_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q1 + RT_TIC_Q2 + RT_TIC_Q3) |>
    ggplot(aes(x = `rtOverTicQuantile.75%`, y = RT_TIC_Q3)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 75% (MS1, MsQuality)") + 
        ylab("RT_TIC_Q3 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_tic_q3_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q1 + RT_TIC_Q2 + RT_TIC_Q3) |>
    ggplot(aes(x = `rtOverTicQuantile.75%`, y = RT_TIC_Q3)) +
        stat_density_2d() +
        xlab("rtOverTicQuantile 75% (MS2, MsQuality)") + 
        ylab("RT_TIC_Q3 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

ggpubr::ggarrange(gg_rt_tic_q1_msLevel1, gg_rt_tic_q2_msLevel1, gg_rt_tic_q3_msLevel1,
    gg_rt_tic_q1_msLevel2, gg_rt_tic_q2_msLevel2, gg_rt_tic_q3_msLevel2,
    ncol = 3, nrow = 2, labels = "AUTO")
```

```{r rtTicOverQuantile_Q1Q2Q3_in_one_plot, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: RT over TIC quantile. The metrics are calculated from MS1 and MS2 spectra."}
## combine Q1, Q2, and Q3 in one plot per msLevel
gg_rt_tic_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q2 + RT_TIC_Q3) |>
    select(rowname,  `rtOverTicQuantile.25%`, `rtOverTicQuantile.50%`, 
        `rtOverTicQuantile.75%`, RT_TIC_Q1, RT_TIC_Q2, RT_TIC_Q3) |>
    rename("Q1_rtOverTicQuantile" = `rtOverTicQuantile.25%`,
        "Q2_rtOverTicQuantile" = `rtOverTicQuantile.50%`,
        "Q3_rtOverTicQuantile" = `rtOverTicQuantile.75%`,
        "Q1_RTTIC" = RT_TIC_Q1, "Q2_RTTIC" = RT_TIC_Q2, 
        "Q3_RTTIC" = RT_TIC_Q3) |>
    pivot_longer(-rowname, names_to = c("Quantile", ".value"), names_sep = "_") |>
    ggplot(aes(x = rtOverTicQuantile, y = RTTIC)) +
        stat_density_2d(aes(color = Quantile)) +
        xlab("rtOverTicQuantile (MS1, MsQuality)") + 
        ylab("RT_TIC (QuaMeter)") +
        xlim(0, 1) + ylim(0, 1) +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_tic_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q2 + RT_TIC_Q3) |>
    select(rowname,  `rtOverTicQuantile.25%`, `rtOverTicQuantile.50%`, 
        `rtOverTicQuantile.75%`, RT_TIC_Q1, RT_TIC_Q2, RT_TIC_Q3) |>
    rename("Q1_rtOverTicQuantile" = `rtOverTicQuantile.25%`,
        "Q2_rtOverTicQuantile" = `rtOverTicQuantile.50%`,
        "Q3_rtOverTicQuantile" = `rtOverTicQuantile.75%`,
        "Q1_RTTIC" = RT_TIC_Q1, "Q2_RTTIC" = RT_TIC_Q2, 
        "Q3_RTTIC" = RT_TIC_Q3) |>
    pivot_longer(-rowname, names_to = c("Quantile", ".value"), names_sep = "_") |>
    ggplot(aes(x = rtOverTicQuantile, y = RTTIC)) +
        stat_density_2d(aes(color = Quantile)) +
        xlab("rtOverTicQuantile (MS2, MsQuality)") + 
        ylab("RT_TIC (QuaMeter)") +
        xlim(0, 1) + ylim(0, 1) +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

ggpubr::ggarrange(gg_rt_tic_msLevel1, gg_rt_tic_msLevel1,
    ncol = 2, nrow = 1, labels = "AUTO")
```

```{r rtOverMsQuarters, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: RT over MS quarters. The metrics are calculated from MS1 and MS2 spectra."}
##
## rtOverMsQuarters vs RT_MS_Q1
gg_rt_msQuarters_q1_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter1`, y = RT_MS_Q1)) +
        geom_point(alpha = 0.2) +
        xlab("Q1 (MsQuality)") + 
        ylab("Q1 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

## rtOverMsQuarters vs RT_MS_Q2
gg_rt_msQuarters_q2_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MS_Q2 = RT_MS_Q1 + RT_MS_Q2) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter2`, y = RT_MS_Q2)) +
        geom_point(alpha = 0.2) +
        xlab("Q2 (MsQuality)") + 
        ylab("Q2 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_msQuarters_q3_msLevel1 <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MS_Q3 = RT_MS_Q1 + RT_MS_Q2 + RT_MS_Q3) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter3`, y = RT_MS_Q3)) +
        geom_point(alpha = 0.2) +
        xlab("Q3 (MsQuality)") + 
        ylab("Q3 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

## 
## rtOverMsQuarters vs RT_MSMS_Q1
gg_rt_msQuarters_q1_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter1`, y = RT_MSMS_Q1)) +
        geom_point(alpha = 0.2) +
        xlab("Q1 (MsQuality)") + 
        ylab("Q1 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

## rtOverMsQuarters vs RT_MSMS_Q2
gg_rt_msQuarters_q2_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MSMS_Q2 = RT_MSMS_Q1 + RT_MSMS_Q2) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter2`, y = RT_MSMS_Q2)) +
        geom_point(alpha = 0.2) +
        xlab("Q2 (MsQuality)") + 
        ylab("Q2 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_msQuarters_q3_msLevel2 <- .metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MSMS_Q3 = RT_MSMS_Q1 + RT_MSMS_Q2 + RT_MSMS_Q3) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter3`, y = RT_MSMS_Q3)) +
        geom_point(alpha = 0.2) +
        xlab("Q3 (MsQuality)") + 
        ylab("Q3 (QuaMeter)") +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

row1 <- annotate_figure(
    ggarrange(gg_rt_msQuarters_q1_msLevel1, gg_rt_msQuarters_q2_msLevel1, 
    gg_rt_msQuarters_q3_msLevel1, ncol = 3, nrow = 1, labels = c("A", "B", "C")), left = "MS1")
row2 <- annotate_figure(
    ggarrange(gg_rt_msQuarters_q1_msLevel2, gg_rt_msQuarters_q2_msLevel2, 
    gg_rt_msQuarters_q3_msLevel2, ncol = 3, nrow = 1, labels = c("D", "E", "F")), left = "MS2")
p <- ggarrange(row1, row2, ncol = 1, nrow = 2)

ggpubr::annotate_figure(p, top = ggpubr::text_grob("RT over MS Quarters"))
```

```{r msSignal10xChange, fig.cap = "Scatterplot between quality metrics calculated by MsQuality and QuaMeter: MS Signal >10x Change (jump). The metrics are only calculated from MS1 spectra."}
##
## msSignal10xChange vs. IS-1A (jumps)
p <- .metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(IS_1A = as.numeric(IS_1A)) |>
    ggplot(aes(x = msSignal10xChange, y = IS_1A)) +
        geom_point(alpha = 0.2) +
        xlab(" MsQuality") + 
        ylab(" QuaMeter") +
        theme_classic() +
        xlim(0, 1500) + ylim(0, 1500) +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

ggpubr::annotate_figure(p, top = ggpubr::text_grob("MS Signal >10x Change (jump)"))
```

These plots indicate that the QuaMeter and MsQuality generally compute
similar values. 

## Calculate Peak RAM Used

Similar to the above-mentioned analysis using the flow injection
analysis, an important aspect, especially when dealing with large amount of data, 
is scalability and performance when computing the quality metric. 
Again, the total and peakRAM is monitored via `peakRAM`.

```{r df_ram_Amidan2014, eval = FALSE, echo = TRUE}
.path <- "/scratch/naake/Amidan2014"
fls <- dir(.path, full.names = TRUE, recursive = TRUE, pattern = "mzML") |>
    unique()

.metrics <- c("rtDuration", "rtOverTicQuantile", "rtOverMsQuarters", 
    "ticQuartileToQuartileLogRatio", "numberSpectra", 
    "medianPrecursorMz", "rtIqr", "rtIqrRate", "areaUnderTic", 
    "areaUnderTicRtQuantiles", "medianTicRtIqr", "medianTicOfRtRange",
    "mzAcquisitionRange", "rtAcquisitionRange", "precursorIntensityRange", 
    "precursorIntensityQuartiles", "precursorIntensityMean", 
    "precursorIntensitySd", "msSignal10xChange", "ratioCharge1over2", 
    "ratioCharge3over2", "ratioCharge4over2", "meanCharge", 
    "medianCharge")

df_ram <- peakRAM(
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 1))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 2))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 4))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 8))
	},
    function() {
		bplapply(fls, function(fls_i) {
			calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
				metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 16))
	}
)
```

```{r df_ram_saveRDS_Amidan2014, eval = FALSE, echo = FALSE}
## save the data.frame as RDS object
saveRDS(df_ram, file = "Amidan2014/Amidan2014_df_ram.RDS")
```


```{r df_mb_readRDS_Amidan2014, eval = TRUE, echo = FALSE}
df_ram <- readRDS("Amidan2014/Amidan2014_df_ram.RDS")
```

```{r df_ram_plot_Amidan2014, fig.cap = "Total and peak RAM for the calculation of quality metrics of the data set of Amidan2014 et al. (2022) under parallelization (1, 2, 4, 8, and 16 workers)."}
df_ram[["Function_Call"]] <- c(1, 2, 4, 8, 16)

df_ram <- df_ram |> 
    pivot_longer(cols = 2:ncol(df_ram), names_to = "parameter") |>
    mutate(parameter = case_when(
        parameter == "Elapsed_Time_sec" ~ "elapsed time (sec)",
        parameter == "Total_RAM_Used_MiB" ~ "used total RAM",
        parameter == "Peak_RAM_Used_MiB" ~ "used peak RAM"))

## visualize the elapsed time per amount of used workers
df_ram |>
    filter(parameter != "elapsed time (sec)") |>
    ggplot(aes(x = Function_Call, y = value)) +
        geom_point(size = 3) +
        geom_line() +
        xlab("number of workers") + ylab("MiB") +
        scale_x_continuous(breaks = df_ram[["Function_Call"]]) +
        facet_wrap(~ parameter, scales = "free_y") +
        scale_y_log10() +
        theme_bw() +
        theme(axis.title = element_text(size = 16), 
            axis.text = element_text(size = 14),
            panel.grid = element_blank())
```

We observe here that overhead is controlled when running on 2 or more morkers 
and, in these cases, the peak and total RAM does not scale with the number of cores. For this application, it is recommended to run the jobs within a 
cluster/server environment.

Similarly to above, we measure the time it takes to calculate the  
quality metrics under parallelization of the tasks on 
1, 2, 4, 8, and 16 workers using the `microbenchmark` package. For computational
reasons we limit the calculation on the first 500 mzML files.

```{r df_mb_Amidan2014, echo = TRUE, eval = FALSE}
library("microbenchmark")
.path <- "/scratch/naake/Amidan2014"
fls <- dir(.path, full.names = TRUE, recursive = TRUE, pattern = "mzML") |>
    unique()
fls <- fls[1:1000]

.metrics <- c("rtDuration", "rtOverTicQuantile", "rtOverMsQuarters", 
    "ticQuartileToQuartileLogRatio", "numberSpectra", 
    "medianPrecursorMz", "rtIqr", "rtIqrRate", "areaUnderTic", 
    "areaUnderTicRtQuantiles", "medianTicRtIqr", "medianTicOfRtRange",
    "mzAcquisitionRange", "rtAcquisitionRange", "precursorIntensityRange", 
    "precursorIntensityQuartiles", "precursorIntensityMean", 
    "precursorIntensitySd", "msSignal10xChange", "ratioCharge1over2", 
    "ratioCharge3over2", "ratioCharge4over2", "meanCharge", 
    "medianCharge")

df_mb <- microbenchmark(
    workers_1 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 1)),
    workers_2 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 2)),
    workers_4 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 4)),
    workers_8 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 8)),
    workers_16 = bplapply(fls, function(fls_i) {
		calculateMetricsFromSpectra(spectra = sps[sps$dataOrigin == fls_i, ], 
			metrics = .metrics)}, BPPARAM = MulticoreParam(workers = 16)), 
	times = 32L, control = list(warmup = 2), check = "equal"
)
```

```{r df_mb_saveRDS_Amidan2014, eval = FALSE, echo = FALSE}
## save the data.frame
saveRDS(df_mb, file = "Amidan2014/Amidan2014_df_mb.RDS")
```

```{r df_mb_readRDS_Amidan2014, eval = TRUE, echo = FALSE}
df_mb <- readRDS(file = "Amidan2014/Amidan2014_df_mb.RDS")
```

```{r df_mb_plot_Amidan2014, echo = TRUE, eval = TRUE, fig.cap = "Execution time for the calculation of quality metrics of the data set of Amidan et al. (2014) under parallelization (1, 2, 4, 8, and 16 workers)."}
## convert from nano seconds to seconds
df_mb[["time"]] <- df_mb[["time"]] / 10e9
df_mb <- df_mb |>
    as.data.frame() |>
    mutate(expr = case_when(expr == "workers_1" ~ 1,
        expr == "workers_2" ~ 2, expr == "workers_4" ~ 4,
        expr == "workers_8" ~ 8, expr == "workers_16" ~ 16))

## visualize the elapsed time per amount of used workers
ggplot(df_mb) +
    geom_jitter(aes(x = expr, y = time, col = as.factor(expr)), alpha = 0.2) +
    geom_violin(aes(x = expr, y = time, group = expr, fill = NA), fill = alpha("white", 0)) +
    xlab("number of workers") + ylab("elapsed time (sec)") +
    scale_x_continuous(breaks = df_mb[["expr"]]) +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    theme_classic() +
    ylim(50, 800) +
    theme(axis.title = element_text(size = 16), 
            axis.text = element_text(size = 14)) +
    theme(legend.position = "none")
```

The `microbenchmark` package was used to accurately measure the performance 
improvements achieved by parallelization. By parallelizing the calculation of 
the quality metrics across multiple workers, it is possible to significantly 
reduce the execution time



\newpage

# References

@article{
    name = {},
    title = {},
    journal = {},
    volume = {},
    issue = {},
    year = {},
    pages = {},
    doi = {}
}

@article{Amidan2014,
    name = {Amidan, B.G. and Orton, D.J. and Lamarche, B.L. and Monroe, M.E. and
        Moore, R.J. and Venzin, A.M. and Smith, R.D. and Sego, L.H. and 
        Tardiff, M.F. and Payne, S.H.},
    title = {Signatures for mass spectrometry data quality},
    journal = {Proteome Research},
    volume = {13},
    issue = {4},
    year = {2014},
    pages = {2215--2222},
    doi = {10.1021/pr401143e}
}

@article{Bereman2015,
    name = {Bereman, M.S.},
    title = {Tools for Monitoring System Suitability in LC MS/MS Centric 
        Proteomic Experiments},
    journal = {Proteomics},
    volume = {15},
    year = {2015},
    pages = {891--902},
    doi = {10.1002/pmic.201400373}
}

@article{Cherkaoui2022,
    name = {Cherkaoui, S. and Durot, S. and Bradley, J. and Critchlow, S. and
        Dubuis, S. and Masiero, M.M. and Wegmann, R. and Snijder, B. and 
        Othman, A. and Bendtsen, C. and Zamboni, N.},
    title = {A functional analysis of 180 cancer cell lines reveals 
        conserved intrinsic metabolic programs},
    journal = {Molecular Systems Biology},
    volume = {18},
    year = {2022},
    pages = {e11033},
    doi = {10.15252/msb.202211033}
}

@article{Koecher2011,
    name = {K\"ocher, T. and Pichler, P. and Swart, R. and Mechtler, K.},
    title = {Quality control in LC-MS/MS},
    journal = {Proteomics and Systems Biology},
    volume = {11},
    issue = {6},
    year = {2011},
    pages = {1026--1030},
    doi = {10.1002/pmic.201000578}
}

@article{Paulovich2010,
    name = {Paulovich, A.G. and Billheimer, D. and Ham, A.-J. L. and 
        Vega-Montoto, L. and Rudnick, P.A. and Tabb, D.L. and Wang, P. and 
        Blackman, D.M. and Bunk, D.M. and Cardasis, H.L. and Clauser, K.R. and
        Kinsinger, C.R. and Schilling, B. and Tegeler, T.J. and Variyath, A.M. 
        and Wang, M. and Whiteaker, J.R. and Zimmerman, L.J. and Fenyo, D. and 
        Carr, S.A. and Fisher, S.J. and Gibson, B.W. and Mesri, M. and Neubert, 
        T.A. and Regnier, F.E. and Rodriguez, H. and Spiegelman, C. and 
        Stein, S.E. and Tempst, P. and Liebler, D.C.},
    title = {Interlaboratory Study Characterizing a Yeast Performance Standard 
        for Benchmarking LC-MS Platform Performance},
    journal = {Molecular & Cellular Proteomics},
    volume = {9},
    issue = {2},
    year = {2010},
    pages = {242--254},
    doi = {10.1074/mcp.M900222-MCP200}
}

@article{Ma2012,
    name = {Ma, Z.-Q. and Polzin, K.O. and Dasari, S. and Chambers, M.C. and 
        Schilling, B. and Gibson, B.W. and Tran, B.Q. and Vega-Montoto, L. and
        Liebler, D.C. and Tabb, D.L.},
    title = {QuaMeter: Multivendor Performance Metrics for LC–MS/MS Proteomics 
        Instrumentation},
    journal = {Analytical Chemistry},
    volume = {84},
    issue = {14},
    year = {2012},
    pages = {5845-5850},
    doi = {10.1021/ac300629p}
}

@article{Rainer2022,
    name = {Rainer, J. and Vicini, A. and Salzer, L. and Stanstrup, J. and
        Badia, J.M. and Neumann, S. and Stravs, M.A. Hernandes, V.V. and
        Gatto, L. and Gibb, S. and Witting, M.},
    title = {A Modular and Expandable Ecosystem for Metabolomics Data Annotation 
        in R},
    journal = {Metabolites},
    volume = {12},
    issue = {2},
    year = {2022},
    pages = {173},
    doi = {10.3390/metabo12020173}
}

