---
title: "MsQuality – an interoperable open-source package for the calculation of standardized quality metrics of mass spectrometry data"
subtitle: "Supplementary Data"
author:
    - Thomas Naake^[Genome Biology Unit, European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg, Germany]
    - Johannes Rainer^[Institute for Biomedicine (Affiliated Institute of the University of Lübeck), Eurac Research, Viale Druso 1, 39100 Bolzano, Italy]
    - Wolfgang Huber^[Genome Biology Unit, European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg, Germany]
package: MsQuality

bibliography: document.bib
fontsize: 12pt
output:
  bookdown::pdf_document2:
    includes:
      in_header: "preamble.tex"
    extra_dependencies: ["natbib", "float", "lscape", "subfig", "graphicx"]
    fig_caption: true
    keep_tex: true
    citation_package: natbib
    toc_depth: 3
    number_sections: true
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

\beginsupplement

\newpage

This is the Supplementary Information for the publication "MsQuality - an 
interoperable open-source package for the calculation of standardized quality
metrics of mass spectrometry data". It demonstrates  
functionality of the `MsQuality` package on two example analysis 
workflows using the data sets of @Cherkaoui2022 and @Amidan2014.

The `MsQuality` package calculates low-level quality metrics that only require 
minimal information about the mass spectrometry data: retention time, m/z 
values, and associated intensities. The list of quality metrics provided by the
mzQC framework ([hupo-psi.github.io/mzQC](hupo-psi.github.io/mzQC)) 
is extensive, also including metrics that depend on  
higher level information which might not be readily accessible from .raw or 
.mzML files, such as pump pressure mean, or that rely on alignment results, 
like retention time mean shift, signal-to-noise ratio, precursor errors (ppm). 
Such metrics are currently not implemented in `MsQuality`. 

The `MsQuality` package relies on the `Spectra` package for data import and
representation. Quality metrics are calculated from the information
in a `Spectra` object. The `dataOrigin` variable is
used to distinguish between the MS data from different measurements/files.
Section \ref{sec:env} loads these and other packages into the environment
of the `R` session in order to run all analyses.

In subsequent sections of this document the quality of two 
data sets will be analyzed:

- Section \ref{sec:Cherkaoui2022}: the @Cherkaoui2022 data set is a mass spectrometry (MS) metabolomics data set of 
180 cancer cell lines obtained via flow injection analysis 
(TOF, negative ionization mode). The data set comprises a total of 1397 
measurements.

- In Section \ref{sec:Amidan2014}: the @Amidan2014 data set consists 
of 3431 LC-MS measurements of a single QC sample 
(whole cell lysate of *Shewanella oneidensis*). The QC sample was measured on 
Exactive, LTQ IonTrap, LTQ Orbitrap, and Velos Orbitrap instruments.

We note that these quality metrics are indicative, but by themselves might not 
be sufficient for data quality control decision-making, such as removing 
low-quality measurements, which might require additional
consideration of more advanced analytics, such as those provided by `MatrixQCvis`
[@Naake2022]. As stated previously [@Bittremieux2017], the utility of  QC metrics 
depend on the type of the sample, e.g., on whether a single peptide or a complex 
lysate of proteins is analyzed [@Bereman2015,@Koecher2011,@Paulovich2010]. 

In this document, we will 
- create `Spectra` objects from the raw data of the two datasets, 
- calculate the quality metrics on these data sets, 
- visualize some of the metrics, and 
- assess performance and scalability of the implemented algorithms using the `microbenchmark` package.

Due to journal's publication format, this document presents static plots.
Note that the `MsQuality` package also includes an interactive shiny application to 
interactively navigate quality metrics, with plots based on the plotly framework. 
For reproducibility, we provide the source .Rmd file in the accompanying
[GitHub](https://www.github.com/tnaake/MsQuality_manuscript/) repository.

A list of the attached pacakges can be found in Section \ref{sec:Session Info}.
We will indicate which parts of this document are reproducible.

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, include=FALSE, echo=FALSE, cache=FALSE}
library("knitr")
knitr::opts_chunk$set(stop_on_error = 1L, fig.pos = "ht")
suppressPackageStartupMessages(library("Spectra"))
suppressPackageStartupMessages(library("mzR"))
suppressPackageStartupMessages(library("MsQuality"))
suppressPackageStartupMessages(library("microbenchmark"))

#knitr::opts_knit$set(root.dir = "")
```

# Preparation of the environment \label{sec:env}

This analysis uses functions from multiple `R` packages, including `Spectra` 
for representing mass spectrometry spectral data and `MsQuality` for calculating
quality metrics. Other packages are required for data visualization 
(`ggplot2`, `ggbeeswarm`, `ggpubr`), data wrangling (`dplyr`, `readxl`, 
`stringr`, `tibble`, `tidyr`), and performance and scalability analysis 
(`microbenchmark`). Before starting the analysis, ensure to load these packages.

```{r load_package, echo = TRUE, eval = TRUE, warning=FALSE, message=FALSE}
## load packages for visualization
library("ggplot2")
library("ggbeeswarm")
library("ggpubr")

## load packages for data wrangling
library("dplyr")
library("readxl")
library("stringr")
library("tibble")
library("tidyr")

## load packages for performance and scalability analysis
library("microbenchmark")

## load packages for storing spectral data and calculating quality metrics
library("Spectra")
library("MsQuality")
```

\newpage

# @Cherkaoui2022: A functional analysis of 180 cancer cell lines reveals conserved intrinsic metabolic programs \label{sec:Cherkaoui2022}

The .mzML files were downloaded from the MassIVE database (accession number 
MSV000087155, available at https://massive.ucsd.edu/) via
ftp://massive.ucsd.edu/MSV000087155. 

We use the [BiocFileCache]() package from Bioconductor to download and cache the
mzML files locally. To this end we first determine below the full file names of
all mzML files of this data set.

All parts in the section 
*@Cherkaoui2022: A functional analysis of 180 cancer cell lines reveals conserved intrinsic metabolic programs*
are reproducible except the parallelization steps in the subsection
**Performance under parallelization**. For these steps precalculated objects
are loaded to the environment.


```{r set_directory_for}
url <- "ftp://massive.ucsd.edu/MSV000087155/ccms_peak/New_mzMLFinal/"
library(curl)
file_list <- readLines(curl(url, "r"))
ftp_files <- strsplit(file_list, " +")
ftp_files <- vapply(
    ftp_files, function(z) paste0(tail(z, 2), collapse = " "), character(1))
```

With the file names available we next create a *BiocFileCache* instance adding
the files. The *BiocFileCache* will take care of downloading files that are not
already available in the local cache thus preventing unneeded downloads.

```{r biocFileCache_cherkaoui2022, message = FALSE}
library(BiocFileCache)
## every additional result should be saved in there
cherkaoui <- BiocFileCache("../Cherkaoui2022", ask = FALSE)
path <- bfcrpath(cherkaoui, paste0(url, curl_escape(ftp_files)))
```

These downloaded mzML files can however not be directly loaded because they are
not fully compliant with the open mzML standard file format (internal references
to instrumentation configuration are missing). We thus need to process all files
to remove these incompatible lines from each mzML file. This needs to be done
(once) using the below unix shell commands that should be executed in the folder
containing the downloaded files.

```{bash}
cd ../Cherkaoui2022
sed -i 's/<run defaultInstrumentConfigurationRef=.*/<run/g' *.mzML
sed -i '/^<scanWindowList/,/^<\/scanWindowList/d' *.mzML
```


## Instantiation of the Spectra object

In the subsequent analysis, a `Spectra` object is instantiated. The operations 
were executed within a (high-performance) computing environment (31 cores, 
64 GB RAM pool for all cores).

```{r create_spectra_Cherkaoui, message = FALSE, cache = TRUE}
## create the Spectra object
sps <- Spectra(path, backend = MsBackendMzR())
```


## Calculate the metrics via `MsQuality`

`MsQuality` uses the `Spectra` class for storing the spectral data. In 
this particular case, where the spectral data was obtained via flow injection 
analysis, metrics that incorporate retention time information are not 
relevant and the analysis will only focus on the three metrics

 - `numberSpectra`, **Number of MS1 spectra** (QC:4000059), “The number of MS1 events in the run.” [PSI:QC];
 - `areaUnderTic`, **Area under TIC** (QC:4000077), “The area under the total ion chromatogram.” [PSI:QC];
 - `mzAcquisitionRange`, **m/z acquisition range** (QC:4000138), “Upper and lower limit of m/z values at which spectra are recorded.” [PSI:QC].


The metrics are calculated using the function 
`calculateMetricsFromSpectra`, which takes as input the `Spectra` object, 
`sps`, and the above-defined metrics.
Optional parameters can also be passed to this function for further control of 
the calculation, such as `msLevel` for cases where multiple mass spectra levels 
are present in the `Spectra` object. It is unnecessary to specify `msLevel` 
in the current context since only MS1 level spectra are stored in the 
`Spectra` object.


```{r calculateMetrics_Cherkaoui2022, cache = TRUE}
metrics <- c("numberSpectra", "areaUnderTic", "mzAcquisitionRange")

metrics_sps <- calculateMetricsFromSpectra(spectra = sps, 
    metrics = metrics)
```


## Visualization

We next visualize the three quality metrics using the `ggplot2` package. We
include also information from the original study @Cherkaoui2022 in particular
which of the files were included in the final analysis. The results of the study
are available from this resource: https://doi.org/10.3929/ethz-b-000511784 . 
We first download and cache the *PrimaryAnalysis.zip* archive that contains all
results, unzip it to a temporary folder and import the
*metabolomics_180CCL.xlsx* file.

```{r}
l <- paste0("https://www.research-collection.ethz.ch/bitstream/handle/",
            "20.500.11850/511784/PrimaryAnalysis.zip?sequence=1&isAllowed=y")
arch <- bfcrpath(cherkaoui, l)
unzip(zipfile = arch, files = "PrimaryAnalysis/Metabolomics_180CCL.xlsx",
      exdir = tempdir())
measurements <- read_xlsx(
    file.path(tempdir(), "PrimaryAnalysis/Metabolomics_180CCL.xlsx"),
    sheet = "injections")
```

From this excel sheet we retrieve the information whether a measurements was
analyzed or ecluded and add this information to the `metrics_sps` object with
the quality information.

We then create a Figure to compare the differences in quality metrics 
between the analyzed and excluded measurements (Figure
\@ref(fig:metrics-cherkaoui2022)).

```{r echo = FALSE}
## reshape the metrics into long format
metrics_sps <- metrics_sps |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble() |>
    pivot_longer(cols = 2:(ncol(metrics_sps) + 1))

## for visualization purposes truncate the dataOrigin, the rowname entry
## will contain 
metrics_sps[["rowname"]] <- metrics_sps[["rowname"]] |>
    strsplit(split = "_2016") |>
    lapply(FUN = function(names_i) paste0("2016", names_i[[2]])) |> 
    unlist() |>
    strsplit(split = "%") |>
    lapply(FUN = function(names_i) names_i[[1]]) |>
    unlist()

## add information if the measurement was analyzed or excluded
## (information stored in measurements)
metrics_sps <- metrics_sps |> 
    mutate(dsCode = sapply(
        strsplit(metrics_sps$rowname, split = "_Batch"), "[", 1)) |>
    mutate(analyzed = ifelse(dsCode %in% measurements[["dsCode"]], "analyzed", "excluded"))
```

```{r metrics-cherkaoui2022, fig.cap = "Quality metrics for data set of @Cherkaoui2022 stratified by information if the measurement was analyzed (`yes`) or excluded (`no`). (a) Area under the TIC (`areaUnderTic`). (b) Minimum values of the m/z acquisition range (`mzAcquisitionRange.min`). (c) maximum values of the m/z acquisition range `mzAcquisitionRange.max`). A.U.: arbitrary units.", echo = FALSE}
gg_areaUnderTic <- metrics_sps |>
    filter(name == "areaUnderTic") |>
    ggplot() +
        geom_beeswarm(aes(x = analyzed, y = value, col = analyzed), cex = 0.8, size = 0.001, method = "compactswarm") +
        #geom_violin(aes(x = analyzed, y = value, col = analyzed)) +
        #geom_point(aes(x = analyzed, y = value, col = analyzed), 
        #    position = position_jitter(width = 0.2, height = 0), alpha = 0.2) +
        ylab("area under TIC (A.U.)") +
        theme_classic() +
        theme(axis.title.x = element_blank(), legend.position = "none")

gg_mzRange_min <- metrics_sps |>
    filter(name == "mzAcquisitionRange.min") |>
    ggplot() +
        geom_beeswarm(aes(x = analyzed, y = value, col = analyzed), cex = 0.05, size = 0.001, method = "compactswarm") +
        #geom_violin(aes(x = analyzed, y = value, col = analyzed)) +
        #geom_point(aes(x = analyzed, y = value, col = analyzed), 
        #    position = position_jitter(width = 0.2, height = 0), alpha = 0.2) +
        ylab("minimal m/z value") +
        theme_classic() +
        theme(axis.title.x = element_blank(), legend.position = "none")

gg_mzRange_max <- metrics_sps |>
    filter(name == "mzAcquisitionRange.max") |>
    ggplot() +
        geom_beeswarm(aes(x = analyzed, y = value, col = analyzed), cex = 0.8, size = 0.001, method = "compactswarm") +
        #geom_violin(aes(x = analyzed, y = value, col = analyzed)) +
        #geom_point(aes(x = analyzed, y = value, col = analyzed), 
        #    position = position_jitter(width = 0.2, height = 0), alpha = 0.2) +
        ylab("maximal m/z value") +
        theme_classic() +
        theme(axis.title.x = element_blank(), legend.position = "none")

ggarrange(gg_areaUnderTic, ggplot() + theme_void(), 
    gg_mzRange_min, gg_mzRange_max,
    ncol = 2, nrow = 2, labels = c("a", "", "b", "c"))
```

Figure \@ref(fig:metrics-cherkaoui2022) demonstrates that the excluded 
measurements show a bimodal distribution of the total ion current (TIC). 
Specifically, some of the excluded measurements have lower total ion current 
(TIC) values, which was already noted in the original publication and was the 
reason for their exclusion from subsequent analysis steps. Figure
\@ref(fig:metrics-cherkaoui2022) a serves as 
a visual confirmation of this statement and aids in understanding the data 
quality of the measurements 
The metrics `mzAcquisitionRange.max` and `mzAcquisitionRange.min` on the other
hand (Figure \@ref(fig:metrics-cherkaoui2022) (b) and (c)) are not informative 
for the decision making on excluding/including the measurements in further 
analysis steps.

## Performance under parallelization

An important aspect, especially when dealing with large amount of data,
is scalability and performance when computing the quality metric. 

By monitoring parallelization, it is possible 
to determine the scalability of the computation and ensure that the 
performance of the analysis remains acceptable as the data size increases.

We measure below the time it takes to evaluate the 
calculation of quality metrics by parallelizing the tasks on
1, 2, 4, 8, and 16 workers using the `microbenchmark` package. This package 
allows for precise measurement of the execution time of `R` expressions by 
repeating the evaluation multiple times and providing detailed summary 
statistics of the execution times. 

```{r df_mb_Cherkaoui2022, echo = TRUE, eval = FALSE}
metrics <- c("numberSpectra", "areaUnderTic", "mzAcquisitionRange")
df_mb <- microbenchmark(calculateMetricsFromSpectra(spectra = sps, 
	    metrics = metrics, BPPARAM = MulticoreParam(workers = 1)),
    workers_2 = calculateMetricsFromSpectra(spectra = sps, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 2)),
    workers_4 = calculateMetricsFromSpectra(spectra = sps, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 4)),
    workers_8 = calculateMetricsFromSpectra(spectra = sps, 
	    metrics = metrics, BPPARAM = MulticoreParam(workers = 8)),
    workers_16 = calculateMetricsFromSpectra(spectra = sps, 
	    metrics = metrics, BPPARAM = MulticoreParam(workers = 16)), 
	times = 110L, control = list(warmup = 10), check = "equal"
)
```

```{r df_mb_saveRDS_Cherkaoui2022, eval = FALSE, echo = FALSE}
## save the data.frame
saveRDS(df_mb, "../Cherkaoui2022/Cherkaoui2022_df_mb.RDS")
```

```{r df_mb_readRDS_Cherkaoui2022, eval = TRUE, echo = FALSE}
df_mb <- readRDS("../Cherkaoui2022/Cherkaoui2022_df_mb.RDS")
```

```{r microbenchmark-cherkaoui2022, echo = FALSE, eval = TRUE, fig.cap = "Execution time for the calculation of quality metrics of the data set of @Cherkaoui2022 under parallelization (1, 2, 4, 8, and 16 workers)."}
## convert from nano seconds to seconds
df_mb[["time"]] <- df_mb[["time"]] / 10e9
df_mb <- df_mb |>
    as.data.frame() |>
    mutate(expr = case_when(expr == "workers_1" ~ 1,
        expr == "workers_2" ~ 2, expr == "workers_4" ~ 4,
        expr == "workers_8" ~ 8, expr == "workers_16" ~ 16))

## visualize the elapsed time per amount of used workers
ggplot(df_mb) +
    geom_jitter(aes(x = expr, y = time, col = as.factor(expr)), 
        alpha = 0.2) +
    geom_violin(aes(x = expr, y = time, group = expr, fill = NA), 
        fill = alpha("white", 0)) +
    xlab("number of workers") + ylab("elapsed time (s)") +
    scale_x_continuous(breaks = df_mb[["expr"]]) +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    theme_classic() + ylim(0, 60) +
    theme(legend.position = "none")
```

By parallelizing the calculation of the 
quality metrics across multiple workers, it is possible to significantly 
reduce the execution time, and the `microbenchmark` package was used to 
accurately measure the performance improvements achieved by parallelization
(Figure \@ref(fig:microbenchmark-cherkaoui2022)). 
The parallelization process can help in 
the management of bigger data sets, and to save valuable time in data analysis.

\newpage

# @Amidan2014: Signatures for mass spectrometry data quality \label{sec:Amidan2014}


The RAW files were downloaded from 

- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000320
  (`1_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000321
  (`2_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000322
  (`3_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000323
  (`4_of_5`),
- ftp://ftp.pride.ebi.ac.uk/pride/data/archive/2013/10/PXD000324
  (`5_of_5`).

Subsequently, the RAW files were converted into `.mzML` files using 
MSConvertGUI (64-bit, v3.0.22015-aadd392) with setting
`peakPicking` to `vendor msLevel=1-`.


The creation of the Figures in the section 
*@Amidan2014: Signatures for mass spectrometry data quality*
are reproducible. Due to long computation time or requirement of an 
environment that enables for parallelization, the creation of the `Spectra` 
object in the subsection **Instantiation of the Spectra object**, the 
calculation of the quality metrics in the subsection 
**Calculate the metrics via `MsQuality`**, and the parallelization steps
in the subsection **Performance under parallelization** are precomputed. 

## Instantiation of the Spectra object

In the subsequent analysis, a `Spectra` object is instantiated. The operations 
were executed within a (high-performance) computing environment (31 cores, 
128 GB RAM pool for all cores), where the .mzML files were stored in the 
directory `Amidan2014`.

```{r create_spectra_Amidan2014, echo = TRUE, eval = FALSE, cache=TRUE}
## read the file with protein intensities
path <- "/scratch/naake/Amidan2014"
fls <- dir(path, full.names = TRUE, recursive = TRUE, pattern = "mzML")

## create the Spectra object
sps <- Spectra(fls, backend = MsBackendMzR())
```

```{r sps_saveRDS_Amidan2014, echo = FALSE, eval = FALSE}
## save the Spectra object as RDS file
saveRDS(sps, file = "../Amidan2014/Amidan2014_sps.RDS")
```



## Calculate the metrics via `MsQuality`


`MsQuality` utilizes `Spectra` objects that store the spectral data. Here,
retention time information was available from the .mzML files and a higher
number of metrics could be calculated:

- *rtDuration*, **RT duration** (QC:4000053), "The retention time duration of 
  the MS run in seconds, similar to the highest scan time minus the lowest 
  scan time." [PSI:QC];
- *rtOverTicQuantiles*, **RT over TIC quantile** (QC:4000054), "The interval 
  when the respective quantile of the TIC accumulates divided by retention time 
  duration. The number of quantiles observed is given by the size of the tuple." 
  [PSI:QC];
- *rtOverMsQuarters*, **MS1 quantiles RT fraction** (QC:4000055), 
  "The interval used for acquisition of the first, second, third, and fourth 
  quarter of all MS1 events divided by RT-Duration." [PSI:QC];
- *rtOverMsQuarters*, **MS2 quantiles RT fraction** (QC:4000055), 
  "The interval used for acquisition of the first, second, third, and fourth 
  quarter of all MS2 events divided by RT-Duration." [PSI:QC];
- *ticQuartileToQuartileLogRatio*, **MS1 TIC-change quartile ratios** 
  (MS:4000057), "The log ratios of successive TIC-change quartiles. The 
  TIC changes are the list of MS1 total ion current (TIC) value changes 
  from one to the next scan, produced when each MS1 TIC is subtracted 
  from the preceding MS1 TIC. The metric's value triplet represents the 
  log ratio of the TIC-change Q2 to Q1, Q3 to Q2, TIC-change-max to Q3" 
  [PSI:MS], `mode = "TIC_change"`;
- *ticQuartileToQuartileLogRatio*, **MS1 TIC quartile ratios** (MS:4000058),
  The log ratios of successive TIC quartiles. The metric's value triplet 
  represents the log ratios of TIC-Q2 to TIC-Q1, TIC-Q3 to TIC-Q2, TIC-max to 
  TIC-Q3." [PSI:MS], `mode = "TIC";
- *numberSpectra*, **Number of MS1 spectra** (QC:4000059), 
  "The number of MS1 events in the run." [PSI:QC]; 
- *numberSpectra*, **Number of MS2 spectra** (QC:4000060), 
  "The number of MS2 events in the run." [PSI:QC];
- *medianPrecursorMz*, **Precursor median m/z for IDs** (QC:4000065), 
  "Median m/z value for all identified peptides (unique ions) after FDR." 
  [PSI:QC];
- *rtIqr*, **Interquartile RT period for peptide identifications** (QC:4000072), 
  "The interquartile retention time period, in seconds, for all peptide 
  identifications over the complete run." [PSI:QC];
- *rtIqrRate*, **Peptide identification rate of the interquartile RT period** 
  (QC:4000073), "The identification rate of peptides for the interquartile 
  retention time period, in peptides per second." [PSI:QC];
- *areaUnderTic*, **Area under TIC** (QC:4000077), "The area under the total ion
  chromatogram." [PSI:QC];
- *areaUnderTicRtQuantiles*, **Area under TIC RT quantiles** (QC:4000078), "The 
  area under the total ion chromatogram of the retention time quantiles. Number 
  of quantiles are given by the n-tuple." [PSI:QC];
- *medianTicRtIqr*, **Median of TIC values in the RT range in which the middle half of peptides are identified** 
  (QC:4000130), "Median of TIC values in the RT range in which half of peptides 
  are identified (RT values of Q1 to Q3 of identifications)" [PSI:QC];
- *medianTicOfRtRange*, **Median of TIC values in the shortest RT range in which half of the peptides are identified** 
  (QC:4000132), "Median of TIC values in the shortest RT range in which half of 
  the peptides are identified"  [PSI:QC] 
- *mzAcquisitionRange*, **m/z acquisition range** (QC:4000138), "Upper and 
  lower limit of m/z values at which spectra are recorded." [PSI:QC];
- *rtAcquisitionRange*, **Retention time acquisition range** (QC:4000139), 
  "Upper and lower limit of time at which spectra are recorded." [PSI:QC];
- *precursorIntensityRange*, **Precursor intensity range** (QC:4000144), 
  "Minimum and maximum precursor intensity recorded." [PSI:QC];
- *precursorIntensityQuartiles*, **Precursor intensity distribution Q1, Q2, Q3**
  (QC:4000167), "From the distribution of precursor intensities, the quartiles 
Q1, Q2, Q3" [PSI:QC];
- *precursorIntensityMean*, **Precursor intensity distribution mean** 
  (QC:4000168), "From the distribution of precursor intensities, the mean." 
  [PSI:QC];
- *precursorIntensitySd*, **Precursor intensity distribution sigma** 
  (QC:4000169), "From the distribution of precursor intensities, the sigma 
  value." [PSI:QC];
- *msSignal10xChange*, **MS1 signal jump (10x) count** (QC:4000172), "The 
  count of MS1 signal jump (spectra sum) by a factor of ten or more (10x) 
  between two subsequent scans" [PSI:QC];
- *RatioCharge1over2*, **Charged peptides ratio 1+ over 2+** (QC:4000174), 
  "Ratio of 1+ peptide count over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge1over2*, **Charged spectra ratio 1+ over 2+** (QC:4000179), 
  "Ratio of 1+ spectra count over 2+ spectra count in all MS2" [PSI:QC];
- *RatioCharge3over2*, **Charged peptides ratio 3+ over 2+** (QC:4000175), 
  "Ratio of 3+ peptide count over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge3over2*, **Charged spectra ratio 3+ over 2+** (QC:4000180), 
  "Ratio of 3+ peptide count over 2+ peptide count in all MS2" [PSI:QC];
- *RatioCharge4over2*, **Charged peptides ratio 4+ over 2+** (QC:4000176), 
  "Ratio of 4+ peptide count  over 2+ peptide count in identified spectra" 
  [PSI:QC];
- *RatioCharge4over2*, **Charged spectra ratio 4+ over 2+** (QC:4000181), 
  "Ratio of 4+ peptide count over 2+ peptide count in all MS2" [PSI:QC];
- *meanCharge*, **Mean charge in identified spectra** (QC:4000177), 
  "Mean charge in identified spectra" [PSI:QC];
- *meanCharge*, **Mean precursor charge in all MS2** (QC:4000182), 
  "Mean precursor charge in all MS2" [PSI:QC];
- *medianCharge*, **Median charge in identified spectra** (QC:4000178), 
  "Median charge in identified spectra" [PSI:QC];
- *medianCharge*, **Median precursor charge in all MS2** (QC:4000183), 
  "Median precursor charge in all MS2" [PSI:QC].

The metrics are calculated using the function 
`calculateMetricsFromSpectra`, which takes as input the `Spectra` object, 
`sps`, and the above-defined metrics.
We calculate the metrics separately for the MS1 (`msLevel 1`) and MS2 spectra
(`msLevel 2`).


```{r, eval = FALSE, cache = TRUE}
metrics <- c("rtDuration", "rtOverTicQuantiles", "rtOverMsQuarters", 
    "ticQuartileToQuartileLogRatio", "numberSpectra", 
    "medianPrecursorMz", "rtIqr", "rtIqrRate", "areaUnderTic", 
    "areaUnderTicRtQuantiles", "medianTicRtIqr", "medianTicOfRtRange",
    "mzAcquisitionRange", "rtAcquisitionRange", "precursorIntensityRange", 
    "precursorIntensityQuartiles", "precursorIntensityMean", 
    "precursorIntensitySd", "msSignal10xChange", "ratioCharge1over2", 
    "ratioCharge3over2", "ratioCharge4over2", "meanCharge", 
    "medianCharge")

metrics_sps_msLevel1 <- calculateMetricsFromSpectra(spectra = sps, 
    metrics = metrics, msLevel = 1L)
metrics_sps_msLevel1 <- calculateMetricsFromSpectra(spectra = sps, 
    metrics = metrics, msLevel = 2L)
```


```{r metrics_sps_saveRDS_Amidan2014, echo = FALSE, eval = FALSE}
saveRDS(metrics_sps_msLevel1, 
    file = "../Amidan2014/Amidan2014_metrics_sps_msLevel1.RDS")
saveRDS(metrics_sps_msLevel2, 
    file = "../Amidan2014/Amidan2014_metrics_sps_msLevel2.RDS")
```

```{r metrics_sps_readRDS_Amidan2014, echo = FALSE, eval = TRUE}
metrics_sps_msLevel1 <- readRDS("../Amidan2014/Amidan2014_metrics_sps_msLevel1.RDS")
metrics_sps_msLevel2 <- readRDS("../Amidan2014/Amidan2014_metrics_sps_msLevel2.RDS")
```

Overall, this function provides a flexible and 
efficient way to analyze large amounts of mass spectrometry data and obtain 
insights on the quality of the data.

## Visualization

In the analysis of the @Amidan2014 study, the quality metrics were
visualized using the `ggplot2` package. The XLS files pr401143e_si_002.xls
and pr401143e_si_003.xls (provided as Supplemental Material of the original
publication) was used to extract information on the measurement quality.
This information was added to the `metrics_sps_msLevel1` and 
`metrics_sps_msLevel2` objects.

The Figures \@ref(fig:areaUnderTic-amidan2014), 
\@ref(fig:areaUnderTicRtQuantiles-amidan2014), and
\@ref(fig:ticQuartileToQuartileLogRatio-amidan2014) were created as examples 
to compare the differences between the low- and high-quality measurements 
for several of the supported quality metrics.

While the metric `rtDuration` (retention time) is a continuous variable,
for visualization purposes we will bin the variable to discrete values and 
will use the measurements over 60 min and 100 min for visualization.

```{r join_metrics_and_metadata, echo = FALSE}
## reshape the metrics into long format
metrics_sps_msLevel1 <- metrics_sps_msLevel1 |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble()
metrics_sps_msLevel2 <- metrics_sps_msLevel2 |>
    as.data.frame() |>
    rownames_to_column(var = "rowname") |>
    as_tibble()

## for visualization purposes truncate the dataOrigin, the rowname entry
## will contain the measurement name
metrics_sps_msLevel1[["rowname"]] <- metrics_sps_msLevel1[["rowname"]] |>
    strsplit(split = "Amidan2014/[1-5]_of_5/", fixed = FALSE) |>
    lapply(FUN = function(names_i) names_i[[2]]) |> 
    unlist() |>
    str_remove(pattern = ".mzML") |>
    make.names()
metrics_sps_msLevel2[["rowname"]] <- metrics_sps_msLevel2[["rowname"]] |>
    strsplit(split = "Amidan2014/[1-5]_of_5/", fixed = FALSE) |>
    lapply(FUN = function(names_i) names_i[[2]]) |> 
    unlist() |>
    str_remove(pattern = ".mzML") |>
    make.names()

## add here information on the classification of measurement (information 
## stored in easurements)
measurements_training <- read_excel("../Amidan2014/pr401143e_si_002.xls",
    sheet = "FileS1_TrainingDataset")
measurements_validation <- read_excel("../Amidan2014/pr401143e_si_003.xls",
    sheet = "FileS2_ValidationDataset")
cols <- intersect(colnames(measurements_training), 
    colnames(measurements_validation))
measurements <- rbind(measurements_training[, cols], measurements_validation[, cols])
measurements[["Dataset"]] <- measurements[["Dataset"]] |> 
    make.names()

metrics_sps_msLevel1 <- inner_join(metrics_sps_msLevel1, measurements, 
    by = c("rowname" = "Dataset"), copy = TRUE)
metrics_sps_msLevel2 <- inner_join(metrics_sps_msLevel2, measurements, 
    by = c("rowname" = "Dataset"), copy = TRUE)

## in the following bin the retention times
metrics_sps_msLevel1$rtDuration_new <- ifelse(
    metrics_sps_msLevel1$rtDuration > 3200 & 
    metrics_sps_msLevel1$rtDuration < 4000, 3600, 
    metrics_sps_msLevel1$rtDuration)
metrics_sps_msLevel1$rtDuration_new <- ifelse(
    metrics_sps_msLevel1$rtDuration_new > 5600 & 
    metrics_sps_msLevel1$rtDuration_new < 6400, 6000, 
    metrics_sps_msLevel1$rtDuration_new)
metrics_sps_msLevel1$rtDuration_new <- ifelse(
    metrics_sps_msLevel1$rtDuration_new > 10400 & 
    metrics_sps_msLevel1$rtDuration_new < 11200, 10800, 
    metrics_sps_msLevel1$rtDuration_new)
metrics_sps_msLevel1$rtDuration_new <- ifelse(
    metrics_sps_msLevel1$rtDuration_new > 14000 & 
    metrics_sps_msLevel1$rtDuration_new < 14800, 14400, 
    metrics_sps_msLevel1$rtDuration_new)
metrics_sps_msLevel1$rtDuration_new <- metrics_sps_msLevel1$rtDuration_new |>
    as.factor()

metrics_sps_msLevel2$rtDuration_new <- ifelse(
    metrics_sps_msLevel2$rtDuration > 3200 & 
    metrics_sps_msLevel2$rtDuration < 4000, 3600, 
    metrics_sps_msLevel2$rtDuration)
metrics_sps_msLevel2$rtDuration_new <- ifelse(
    metrics_sps_msLevel2$rtDuration_new > 5600 & 
    metrics_sps_msLevel2$rtDuration_new < 6400, 6000, 
    metrics_sps_msLevel2$rtDuration_new)
metrics_sps_msLevel2$rtDuration_new <- ifelse(
    metrics_sps_msLevel2$rtDuration_new > 10400 & 
    metrics_sps_msLevel2$rtDuration_new < 11200, 10800, 
    metrics_sps_msLevel2$rtDuration_new)
metrics_sps_msLevel2$rtDuration_new <- ifelse(
    metrics_sps_msLevel2$rtDuration_new > 14000 & 
    metrics_sps_msLevel2$rtDuration_new < 14800, 14400, 
    metrics_sps_msLevel2$rtDuration_new)
metrics_sps_msLevel2$rtDuration_new <- metrics_sps_msLevel2$rtDuration_new |>
    as.factor()
```

```{r helper_function_plot_beeswarm, eval = TRUE, echo = FALSE}
library(patchwork)

plot_beeswarm <- function(metrics, 
    selected_metric = "areaUnderTicRtQuantiles.25%", 
    ylim = c(0, 1.2e12), 
    plot_margin_1 = unit(c(-0.1, 0.5, -0.0, 0), "cm"), 
    plot_margin_2 = unit(c(-0.1, 0.5, -0.0, 0), "cm"), aspect.ratio = 1) {

    .levels <- c("60 min", "100 min") ## 3600, 6000
    metrics <- metrics |>
        mutate(rtDuration_new = as.numeric(as.character(rtDuration_new))) |>
        mutate(rtDuration_new = rtDuration_new / 60) |>
        mutate(rtDuration_new = factor(rtDuration_new, levels = c(60, 100),
                labels = c("60 min", "100 min"))) |>
        filter(rtDuration_new %in% .levels) |> 
        filter(Curated_Quality %in% c("good", "poor")) |>
        filter(!duplicated(rowname)) |>
            select(all_of(c("rowname", "Curated_Quality", "rtDuration_new", 
                selected_metric)))
    
    P <- list()
    
    for (i in seq_along(.levels)) {
        metrics_i <- metrics |>
            filter(rtDuration_new %in% .levels[i]) |>
            pivot_longer(cols = selected_metric) |>
            filter(!is.na(value)) |>
            filter(value != Inf)
        
        P[[i]] <- metrics_i |> 
            ggplot(aes(x = Curated_Quality, y = value, col = Curated_Quality)) +
                geom_beeswarm(aes(x = Curated_Quality, y = value), method = "compactswarm", cex = 0.8, size = 0.001) +
                #geom_point(aes(x = Curated_Quality, y = value), 
                #    position = position_jitter(width = 0.2, height = 0), alpha = 0.05) +
                facet_grid( ~ rtDuration_new, scales = "free_x") +
                #geom_violin(scale = "count") +
                scale_y_continuous(limits = ylim) +
                theme_classic() +
                theme(legend.position = "none", axis.title = element_blank(),
                    plot.margin = plot_margin_1) + 
                theme(aspect.ratio = aspect.ratio)
        
        if (i == 2) 
            P[[i]] <- P[[i]] + 
                theme(axis.text.y = element_blank(), 
                    axis.ticks.y = element_blank(), 
                    plot.margin = plot_margin_2)
        P[[i]]
    }
    
    P[[1]] + P[[2]]
}
```


```{r areaUnderTic-amidan2014, fig.cap = "Quality metrics by `MsQuality`: Area under TIC (`areaUnderTic`). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 measurement run and the data points are displayed as beeswarm plots stratified for high-quality and low-quality measurements as classified in @Amidan2014. (a) Area under TIC for MS1 spectra. (b) Area under TIC for MS2 spectra. A.U.: arbitrary units.", warning = FALSE, echo = FALSE}
xlim <- c(25, 250)
## area under TIC
gg_tic_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
      selected_metric = "areaUnderTic", ylim = c(0, 8e12), 
      plot_margin_1 = unit(c(-15, 0, -15, 0), "cm"), 
      plot_margin_2 = unit(c(-15, 0, -15, 0), "cm"), aspect.ratio = 2)
gg_tic_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
      selected_metric = "areaUnderTic", ylim = c(0, 16e10), 
      plot_margin_1 = unit(c(-15, 0, -15, 0), "cm"), 
      plot_margin_2 = unit(c(-15, 0, -15, 0), "cm"), aspect.ratio = 2) 

p <- ggarrange(gg_tic_msLevel1,  gg_tic_msLevel2, ncol = 2, nrow = 1, 
    labels = "auto")
annotate_figure(p, top = text_grob("area under TIC (A.U.)"))
```

```{r areaUnderTicRtQuantiles-amidan2014, fig.cap = "Quality metrics by `MsQuality`: Area under TIC RT quantiles (`areaUnderTicRtQuantiles`). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 measurement run and the data points are displayed as beeswarm plots stratified for high-quality and low-quality measurements as classified in @Amidan2014. (a) 25% quantile for MS1 spectra. (b) 25% quantile for MS2 spectra. (c) 50% quantile for MS1 spectra. (d) 50% quantile for MS2 spectra.E: 75% quantile for MS1 spectra. (f) 75% quantile for MS2 spectra. (g) 100% quantile for MS1 spectra. (h) 100% quantile for MS2 spectra.", warning = FALSE, echo = FALSE, fig.height = 7.5}

## areaUnderTicRtQuantiles
## areaUnderTicRtQuantiles 25%
gg_tic25_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "areaUnderTicRtQuantiles.25%", ylim = c(0, 1.2e12),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic25_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "areaUnderTicRtQuantiles.25%", ylim = c(0, 3.2e10),
    plot_margin_1 = unit(c(-0.1, -2, -0.0, -2), "cm"), 
    plot_margin_2 = unit(c(-0.1, -2, -0.0, -2), "cm"), aspect.ratio = 1)

## areaUnderTicRtQuantiles 50%
gg_tic50_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "areaUnderTicRtQuantiles.50%", ylim = c(0, 3e12),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic50_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "areaUnderTicRtQuantiles.50%", ylim = c(0, 8e10),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)

## areaUnderTicRtQuantiles 75%
gg_tic75_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "areaUnderTicRtQuantiles.75%", ylim = c(0, 4e12),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic75_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "areaUnderTicRtQuantiles.75%", ylim = c(0, 4.5e10),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)

## areaUnderTicRtQuantiles 100%
gg_tic100_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "areaUnderTicRtQuantiles.100%", ylim = c(0, 12e11),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic100_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "areaUnderTicRtQuantiles.100%", ylim = c(0, 1.2e10),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)

row1 <- ggarrange(
    annotate_figure(ggarrange(gg_tic25_msLevel1, labels = "a"), top = "MS1", 
        left = "25%"),
    annotate_figure(ggarrange(gg_tic25_msLevel2, labels = "b"), top = "MS2"),
    ncol = 2, nrow = 1)
row2 <- ggarrange(
    annotate_figure(ggarrange(gg_tic50_msLevel1, labels = "c"), left = "50%"),
    annotate_figure(ggarrange(gg_tic50_msLevel2, labels = "d")),
    ncol = 2, nrow = 1)
row3 <- ggarrange(
    annotate_figure(ggarrange(gg_tic75_msLevel1, labels = "e"), 
        left = "75%"),
    annotate_figure(ggarrange(gg_tic75_msLevel2, labels = "f")),
    ncol = 2, nrow = 1)
row4 <- ggarrange(
    annotate_figure(ggarrange(gg_tic100_msLevel1, labels = "g"), 
        left = "100%"),
    annotate_figure(ggarrange(gg_tic100_msLevel2, labels = "h")),
    ncol = 2, nrow = 1)
p <- ggarrange(row1, row2, row3, row4, ncol = 1, nrow = 4, align = "v")

annotate_figure(p, top = text_grob("area under TIC RT quantile"))

```

```{r ticQuartileToQuartileLogRatio-amidan2014, fig.cap = "Quality metrics by `MsQuality`: TIC quartile to quartile log ratio (`ticQuartileToQuartileLogRatio`). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 measurement run and the data points are displayed as beeswarm plots stratified for high-quality and low-quality measurements as classified in @Amidan2014. (a) log ratio of quartile 2 to quartile 1 for MS1 spectra. (b) log ratio of quartile 2 to quartile 1 for MS2 spectra. (c) log ratio of quartile 3 to quartile 2 for MS1 spectra. (d) log ratio of quartile 3 to quartile 2 for MS2 spectra. (e) log ratio of quartile 4 to quartile 3 for MS1 spectra.  (f) log ratio of quartile 4 to quartile 3 for MS2 spectra.", warning = FALSE, echo = FALSE, fig.height = 7}

xlim <- c(25, 250)
## ticQuartileToQuartileLogRatio
## ticQuartileToQuartileLogRatio Q2/Q1%
gg_tic_q2_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q2/Q1", ylim = c(0, 3.5),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic_q2_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q2/Q1", ylim = c(0, 3.5),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
  
## ticQuartileToQuartileLogRatio.Q3/Q2
gg_tic_q3_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q3/Q2", ylim = c(0, 3),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic_q3_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q3/Q2", ylim = c(0, 3),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)

## ticQuartileToQuartileLogRatio.Q4/Q3
gg_tic_q4_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q4/Q3", ylim = c(0, 3),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)
gg_tic_q4_msLevel2 <- plot_beeswarm(metrics = metrics_sps_msLevel2, 
    selected_metric = "ticQuartileToQuartileLogRatio.Q4/Q3", ylim = c(1.9, 7.8),
    plot_margin_1 = unit(c(-2.1, -5, -2.0, -5), "cm"), 
    plot_margin_2 = unit(c(-2.1, -5, -2.0, -5), "cm"), aspect.ratio = 1)

row1 <- ggarrange(
    annotate_figure(ggarrange(gg_tic_q2_msLevel1, labels = "a"), top = "MS1"),
    annotate_figure(ggarrange(gg_tic_q2_msLevel2, labels = "b"), top = "MS2"), 
        ncol = 2, nrow = 1) |>
    annotate_figure(left = "Q2/Q1")
row2 <- ggarrange(
    annotate_figure(ggarrange(gg_tic_q3_msLevel1, labels = "c")),
    annotate_figure(ggarrange(gg_tic_q3_msLevel2, labels = "d")), 
        ncol = 2, nrow = 1) |>
    annotate_figure(left = "Q3/Q2")
row3 <- ggarrange(
    annotate_figure(ggarrange(gg_tic_q3_msLevel1, labels = "e")),
    annotate_figure(ggarrange(gg_tic_q3_msLevel2, labels = "f")), 
        ncol = 2, nrow = 1) |>
    annotate_figure(left = "Q4/Q3")

p <- ggarrange(row1, row2, row3, ncol = 1, nrow = 3, align = "v", 
    heights = c(1.1, 1, 1))

annotate_figure(p, top = text_grob("TIC quartile to quartile log ratio"))
```

The Figures \@ref(fig:areaUnderTic-amidan2014) and
\@ref(fig:areaUnderTicRtQuantiles-amidan2014) demonstrate that the 
low-quality measurements (`poor`) have lower total ion current (TIC) values 
compared to high-quality measurements (`good`). 
The Figures serve as a visual check to differences in TIC values and aids in 
understanding the data quality of the measurements. It has to be pointed out that 
a further stratification (e.g. along the instrument type) might be helpful 
to further point out differences between the levels of data quality of the
@Amidan2014 data set.

The Figure \@ref(fig:ticQuartileToQuartileLogRatio-amidan2014) on the other 
hand does not indicate differences between data quality and might 
not be indicate of data quality for the quality issues of the @Amidan2014 data 
set.

### Comparison to QuaMeter metrics

The XLS files also contain pre-calculated `QuaMeter` metrics [@Ma2012] for each of 
the measurements. In the following, we will compare the `QuaMeter`-metrics to the
`MsQuality` metrics to check if `MsQuality` shows concordant results
compared to `QuaMeter`.

```{r rtDurationComparison-amidan2014, fig.cap = "Comparison of quality metrics calculated by `MsQuality` and `QuaMeter`: RT duration (`rtDuration`). The corresponding metric for `QuaMeter` is `RT_Duration` (no specification if the metric was calculated on MS1 and/or MS2 spectra). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities. Brighter areas correspond to high 2D density areas. (a) RT duration for MS1 spectra (`QuaMeter` metric: `RT_Duration`). (b) RT duration for MS2 spectra (`QuaMeter` metric: `RT_Duration`).", echo = FALSE, warning = FALSE}
## rtDuration (MsQuality) vs. RT_Duration (QuaMeter)
gg_rt_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = rtDuration/60, y = RT_Duration/60)) +
        geom_bin_2d(bins = 30) +
        xlab("MsQuality (MS1)") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        xlim(0, 610) + ylim(0, 610) +
        theme(legend.position = "none")

gg_rt_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = rtDuration/60, y = RT_Duration/60)) +
        geom_bin_2d(bins = 30) +
        xlab("MsQuality (MS2)") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        xlim(0, 610) + ylim(0, 610) +
        theme(legend.position = "none")

p <- ggarrange(gg_rt_msLevel1, gg_rt_msLevel2, ncol = 2, nrow = 1, labels = "auto")
annotate_figure(p, top = text_grob("RT duration (min)"))

cor_rtDuration_p <- cor.test(metrics_sps_msLevel1[["rtDuration"]], metrics_sps_msLevel1[["RT_Duration"]], method = "pearson")
cor_rtDuration_s <- cor.test(metrics_sps_msLevel1[["rtDuration"]], metrics_sps_msLevel1[["RT_Duration"]], method = "spearman")
```


```{r numberSpectraComparison-amidan2014, fig.cap = "Comparison of quality metrics calculated by `MsQuality` and `QuaMeter`: Number of spectra (`numberSpectra`). The corresponding metrics for `QuaMeter` are `MS1_Count` and `MS2_Count`. The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities. Brighter areas correspond to high 2D density areas. (a) Number of MS1 spectra (`QuaMeter` metric: `MS1_Count`). (b) Number of MS2 spectra (`QuaMeter` metric: `MS2_Count`).", echo = FALSE, warning = FALSE, message=FALSE}
## 
##  numberSpectra (MsQuality) vs. MS1_Count (QuaMeter)
gg_numberSpectra_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = numberSpectra, y = MS1_Count)) +
        geom_bin_2d(bins = 30) +
        ##geom_point(alpha = 0.2) +
        coord_fixed(ratio = 1) +
        scale_x_continuous(breaks = seq(0, 15100, 5000), limits = c(0, 15100),
            labels = scales::format_format(scientific = FALSE)) +
        scale_y_continuous(breaks = seq(0, 15100, 5000), limits = c(0, 15100),
            labels = scales::format_format(scientific = FALSE)) +
        xlab("MsQuality") + ylab("QuaMeter") +
        theme_classic() +
        theme(legend.position = "none")

##  numberSpectra (MsQuality) vs. MS2_Count (QuaMeter)
gg_numberSpectra_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = numberSpectra, y = MS2_Count)) +
        geom_bin_2d(bins = 30) +
        ##geom_point(alpha = 0.2) +
        xlab("MsQuality") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        scale_x_continuous(breaks = seq(0, 100000, 25000), limits = c(0, 100000),
            labels = scales::format_format(scientific = FALSE)) +
        scale_y_continuous(breaks = seq(0, 100000, 25000), limits = c(0, 100000),
            labels = scales::format_format(scientific = FALSE)) +
        theme_classic() +
        theme(legend.position = "none")


col1 <- annotate_figure(ggarrange(gg_numberSpectra_msLevel1, labels = "a"), 
    top = "MS1") 
col2 <- annotate_figure(ggarrange(gg_numberSpectra_msLevel2, labels = "b"),
    top = "MS2")

p <- ggarrange(col1, col2, ncol = 2, nrow = 1)
annotate_figure(p, top = text_grob("number of spectra"))

## calculate correlation
cor_numberSpectra_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["numberSpectra"]],
    metrics_sps_msLevel1[["MS1_Count"]], method = "pearson")
cor_numberSpectra_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["numberSpectra"]],
    metrics_sps_msLevel1[["MS1_Count"]], method = "spearman")
cor_numberSpectra_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["numberSpectra"]],
    metrics_sps_msLevel2[["MS2_Count"]], method = "pearson")
cor_numberSpectra_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["numberSpectra"]], 
    metrics_sps_msLevel2[["MS2_Count"]], method = "spearman")
```

```{r rtOverTicQuantilesComparison-amidan2014,  fig.cap = "Scatterplot between quality metrics calculated by `MsQuality` and `QuaMeter`: RT over TIC quantile (`rtOverTicQuantiles`). The corresponding metrics for `QuaMeter` are `RT_TIC_Q1`, `RT_TIC_Q2`, and `RT_TIC_Q3` (no specification if these metrices were calculated on MS1 and/or MS2 spectra). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities. Brighter areas correspond to high 2D density areas. (a) 25% quantile for MS1 spectra (`QuaMeter` metric: `RT_TIC_Q1`). (b) 50% quantile for MS1 spectra (`QuaMeter` metric: `RT_TIC_Q2`). (c) 75% quantile for MS1 spectra (`QuaMeter` metric: `RT_TIC_Q3`). (d) 25% quantile for MS2 spectra (`QuaMeter` metric: `RT_TIC_Q1`). (e) 50% quantile for MS2 spectra (`QuaMeter` metric: `RT_TIC_Q2`). (f) 75% quantile for MS2 spectra (`QuaMeter` metric: `RT_TIC_Q3`).", warning = FALSE, echo = FALSE}

## 
##  rtOverTicQuantiles (MsQuality) vs. RT_TIC_Q1 (QuaMeter)
gg_rt_tic_q1_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q1 = RT_TIC_Q1) |>
    ggplot(aes(x = `rtOverTicQuantiles.25%`, y = RT_TIC_Q1)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

gg_rt_tic_q1_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q1 = RT_TIC_Q1) |>
    ggplot(aes(x = `rtOverTicQuantiles.25%`, y = RT_TIC_Q1)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

##  rtOverTicQuantiles (MsQuality) vs. RT_TIC_Q2 (QuaMeter)
gg_rt_tic_q2_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    ggplot(aes(x = `rtOverTicQuantiles.50%`, y = RT_TIC_Q2)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

gg_rt_tic_q2_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    ggplot(aes(x = `rtOverTicQuantiles.50%`, y = RT_TIC_Q2)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

##  rtOverTicQuantiles (MsQuality) vs. RT_TIC_Q3 (QuaMeter)
gg_rt_tic_q3_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q1 + RT_TIC_Q2 + RT_TIC_Q3) |>
    ggplot(aes(x = `rtOverTicQuantiles.75%`, y = RT_TIC_Q3)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

gg_rt_tic_q3_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q1 + RT_TIC_Q2 + RT_TIC_Q3) |>
    ggplot(aes(x = `rtOverTicQuantiles.75%`, y = RT_TIC_Q3)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", axis.title = element_blank(),
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

row1 <- ggarrange(
    annotate_figure(ggarrange(gg_rt_tic_q1_msLevel1, labels = "a"), top = "25%"),
    annotate_figure(ggarrange(gg_rt_tic_q2_msLevel1, labels = "b"), top = "50%"),
    annotate_figure(ggarrange(gg_rt_tic_q3_msLevel1, labels = "c"), top = "75%"),
        ncol = 3, nrow = 1) |>
    annotate_figure(left = "MS1")
row2 <- ggarrange(
    annotate_figure(ggarrange(gg_rt_tic_q1_msLevel2, labels = "d")), 
    annotate_figure(ggarrange(gg_rt_tic_q2_msLevel2, labels = "e")),
    annotate_figure(ggarrange(gg_rt_tic_q3_msLevel2, labels = "f")),
        ncol = 3, nrow = 1) |>
    annotate_figure(left = "MS2")
p <- ggarrange(row1, row2, ncol = 1, nrow = 2)
annotate_figure(p, top = text_grob("RT over TIC Quantiles"))


## calculate correlation
cor_rt_tic_q1_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.25%"]],
    metrics_sps_msLevel1[["RT_TIC_Q1"]], method = "pearson")
cor_rt_tic_q1_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.25%"]],
    metrics_sps_msLevel1[["RT_TIC_Q1"]], method = "spearman")
cor_rt_tic_q1_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.25%"]],
    metrics_sps_msLevel2[["RT_TIC_Q1"]], method = "pearson")
cor_rt_tic_q1_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.25%"]], 
    metrics_sps_msLevel2[["RT_TIC_Q1"]], method = "spearman")

tmp <- metrics_sps_msLevel1[["RT_TIC_Q1"]] + 
    metrics_sps_msLevel1[["RT_TIC_Q2"]]
cor_rt_tic_q2_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.50%"]], tmp, method = "pearson")
cor_rt_tic_q2_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.50%"]], tmp, method = "spearman")
cor_rt_tic_q2_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.50%"]], tmp, method = "pearson")
cor_rt_tic_q2_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.50%"]], tmp , method = "spearman")

tmp <- metrics_sps_msLevel1[["RT_TIC_Q1"]] + 
    metrics_sps_msLevel1[["RT_TIC_Q2"]] +
    metrics_sps_msLevel1[["RT_TIC_Q3"]]
cor_rt_tic_q3_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.75%"]], tmp, method = "pearson")
cor_rt_tic_q3_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverTicQuantiles.75%"]], tmp, method = "spearman")
cor_rt_tic_q3_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.75%"]], tmp, method = "pearson")
cor_rt_tic_q3_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverTicQuantiles.75%"]], tmp, method = "spearman")
```

```{r rtOverTicQuantilesComparisonOnePlot-amidan2014, fig.cap = "Comparison of quality metrics calculated by `MsQuality` and `QuaMeter`: RT over TIC quantiles (`rtOverTicQuantiles`). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities stratified by quantiles 25%, 50%, and 75%. (a) Quantiles for MS1 spectra. (b) Quantile for MS2 spectra.", warning = FALSE, echo = FALSE, fig.show='hide'}
## combine Q1, Q2, and Q3 in one plot per msLevel
gg_rt_tic_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q2 + RT_TIC_Q3) |>
    select(rowname,  `rtOverTicQuantiles.25%`, `rtOverTicQuantiles.50%`, 
        `rtOverTicQuantiles.75%`, RT_TIC_Q1, RT_TIC_Q2, RT_TIC_Q3) |>
    rename("25%_rtOverTicQuantiles" = `rtOverTicQuantiles.25%`,
        "50%_rtOverTicQuantiles" = `rtOverTicQuantiles.50%`,
        "75%_rtOverTicQuantiles" = `rtOverTicQuantiles.75%`,
        "25%_RTTIC" = RT_TIC_Q1, "50%_RTTIC" = RT_TIC_Q2, 
        "75%_RTTIC" = RT_TIC_Q3) |>
    pivot_longer(-rowname, names_to = c("Quantile", ".value"), names_sep = "_") |>
    ggplot(aes(x = rtOverTicQuantiles, y = RTTIC)) +
        stat_density_2d(aes(color = Quantile)) +
        xlab("MsQuality (MS1)") + 
        ylab("QuaMeter") +
        xlim(0, 1) + ylim(0, 1) +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

gg_rt_tic_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q2 + RT_TIC_Q3) |>
    select(rowname,  `rtOverTicQuantiles.25%`, `rtOverTicQuantiles.50%`, 
        `rtOverTicQuantiles.75%`, RT_TIC_Q1, RT_TIC_Q2, RT_TIC_Q3) |>
    rename("25%_rtOverTicQuantiles" = `rtOverTicQuantiles.25%`,
        "50%_rtOverTicQuantiles" = `rtOverTicQuantiles.50%`,
        "75%_rtOverTicQuantiles" = `rtOverTicQuantiles.75%`,
        "25%_RTTIC" = RT_TIC_Q1, "50%_RTTIC" = RT_TIC_Q2, 
        "75%_RTTIC" = RT_TIC_Q3) |>
    pivot_longer(-rowname, names_to = c("Quantile", ".value"), names_sep = "_") |>
    ggplot(aes(x = rtOverTicQuantiles, y = RTTIC)) +
        stat_density_2d(aes(color = Quantile)) +
        xlab("MsQuality (MS2)") + 
        ylab("QuaMeter") +
        xlim(0, 1) + ylim(0, 1) +
        theme_classic() +
        guides(colour = guide_legend(override.aes = list(alpha = 1)))

p <- ggarrange(gg_rt_tic_msLevel1, gg_rt_tic_msLevel2,
    ncol = 2, nrow = 1, labels = "auto")
annotate_figure(p, top = text_grob("RT over TIC quantiles"))
```


```{r rtOverMsQuartersComparison-amidan2014, fig.cap = "Comparison of quality metrics calculated by `MsQuality` and `QuaMeter`: RT over MS quarters (`rtOverMsQuarters`). The `MsQuality` metrics are calculated from MS1 and MS2 spectra. One data point is obtained per MS1 and MS2 spectra and the data points are displayed as 2D densities. Brighter areas correspond to high 2D density areas. (a) Quarter 1 for MS1 spectra. (b) Quarter 2 for MS1 spectra. (c) Quarter 3 for MS1 spectra. (d) Quarter 1 for MS2 spectra. (e) Quarter 2 for MS2 spectra. (f) Quarter 3 for MS2 spectra. Q: quarter.", warning = FALSE, echo = FALSE}

##
## rtOverMsQuarters vs RT_MS_Q1
gg_rt_msQuarters_q1_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter1`, y = RT_MS_Q1)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

## rtOverMsQuarters vs RT_MS_Q2
gg_rt_msQuarters_q2_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MS_Q2 = RT_MS_Q1 + RT_MS_Q2) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter2`, y = RT_MS_Q2)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

## rtOverMsQuarters vs RT_MS_Q3
gg_rt_msQuarters_q3_msLevel1 <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MS_Q3 = RT_MS_Q1 + RT_MS_Q2 + RT_MS_Q3) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter3`, y = RT_MS_Q3)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

##
## rtOverMsQuarters vs RT_MSMS_Q1
gg_rt_msQuarters_q1_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter1`, y = RT_MSMS_Q1)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

## rtOverMsQuarters vs RT_MSMS_Q2
gg_rt_msQuarters_q2_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MSMS_Q2 = RT_MSMS_Q1 + RT_MSMS_Q2) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter2`, y = RT_MSMS_Q2)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

## rtOverMsQuarters vs RT_MSMS_Q3
gg_rt_msQuarters_q3_msLevel2 <- metrics_sps_msLevel2 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_MSMS_Q3 = RT_MSMS_Q1 + RT_MSMS_Q2 + RT_MSMS_Q3) |>
    ggplot(aes(x = `rtOverMsQuarters.Quarter3`, y = RT_MSMS_Q3)) +
        geom_bin_2d(bins = 20) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none")

row1 <- ggarrange(
    annotate_figure(ggarrange(gg_rt_msQuarters_q1_msLevel1, labels = "a"), 
        top = "Q1"),
    annotate_figure(ggarrange(gg_rt_msQuarters_q2_msLevel1, labels = "b"), 
        top = "Q2"),
    annotate_figure(ggarrange(gg_rt_msQuarters_q3_msLevel1, labels = "c"), 
        top = "Q3"),
    ncol = 3, nrow = 1) |>
    annotate_figure(left = "MS1")
row2 <- ggarrange(
    annotate_figure(ggarrange(gg_rt_msQuarters_q1_msLevel2, labels = "d")), 
    annotate_figure(ggarrange(gg_rt_msQuarters_q2_msLevel2, labels = "e")),
    annotate_figure(ggarrange(gg_rt_msQuarters_q3_msLevel2, labels = "f")),
        ncol = 3, nrow = 1) |>
    annotate_figure(left = "MS2")
p <- ggarrange(row1, row2, ncol = 1, nrow = 2)

annotate_figure(p, top = text_grob("RT over MS Quarters"))

## calculate correlation
cor_rt_msQuarters_q1_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter1"]],
    metrics_sps_msLevel1[["RT_MS_Q1"]], method = "pearson")
cor_rt_msQuarters_q1_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter1"]],
    metrics_sps_msLevel1[["RT_MS_Q1"]], method = "spearman")
cor_rt_msQuarters_q1_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter1"]],
    metrics_sps_msLevel2[["RT_MSMS_Q1"]], method = "pearson")
cor_rt_msQuarters_q1_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter1"]], 
    metrics_sps_msLevel2[["RT_MSMS_Q1"]], method = "spearman")

tmp <- metrics_sps_msLevel1[["RT_MS_Q1"]] +
    metrics_sps_msLevel1[["RT_MS_Q2"]]
cor_rt_msQuarters_q2_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter2"]], tmp, method = "pearson")
cor_rt_msQuarters_q2_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter2"]], tmp, method = "spearman")
tmp <- metrics_sps_msLevel1[["RT_MSMS_Q1"]] +
    metrics_sps_msLevel1[["RT_MSMS_Q2"]]
cor_rt_msQuarters_q2_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter2"]], tmp, method = "pearson")
cor_rt_msQuarters_q2_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter2"]], tmp, method = "spearman")

tmp <- metrics_sps_msLevel1[["RT_MS_Q1"]] +
    metrics_sps_msLevel1[["RT_MS_Q2"]] +
    metrics_sps_msLevel1[["RT_MS_Q3"]]
cor_rt_msQuarters_q3_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter3"]], tmp, method = "pearson")
cor_rt_msQuarters_q3_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["rtOverMsQuarters.Quarter3"]], tmp, method = "spearman")
tmp <- metrics_sps_msLevel1[["RT_MSMS_Q1"]] +
    metrics_sps_msLevel1[["RT_MSMS_Q2"]] +
    metrics_sps_msLevel1[["RT_MSMS_Q3"]]
cor_rt_msQuarters_q3_msLevel2_p <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter3"]], tmp, method = "pearson")
cor_rt_msQuarters_q3_msLevel2_s <- cor.test(
    metrics_sps_msLevel2[["rtOverMsQuarters.Quarter3"]], tmp, method = "spearman")
```

```{r msSignal10xChangeComparison-amidan2014, fig.cap = "Comparison of quality metrics calculated by `MsQuality` and `QuaMeter`: MS Signal >10x Change (jump, `msSignal10xChange`). The metrics are only calculated from MS1 spectra. One data point is obtained per MS1 spectra and the data points are displayed as 2D densities. Brighter areas correspond to high 2D density areas.", warning = FALSE, echo = FALSE}

##
## msSignal10xChange vs. IS-1A (jumps)
p <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(IS_1A = as.numeric(IS_1A)) |>
    ggplot(aes(x = msSignal10xChange, y = IS_1A)) +
        geom_bin_2d(bins = 30) +
        xlab("MsQuality") + 
        ylab("QuaMeter") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        xlim(0, 520) + ylim(0, 520) +
        theme(legend.position = "none")

annotate_figure(p, top = text_grob("MS Signal >10x Change (jump)"))

## calculate correlation
cor_msSignal10xChange_msLevel1_p <- cor.test(
    metrics_sps_msLevel1[["msSignal10xChange"]],
    as.numeric(metrics_sps_msLevel1[["IS_1A"]]), method = "pearson")
cor_msSignal10xChange_msLevel1_s <- cor.test(
    metrics_sps_msLevel1[["msSignal10xChange"]],
    as.numeric(metrics_sps_msLevel1[["IS_1A"]]), method = "spearman")
```

The Figures \@ref(fig:rtDurationComparison-amidan2014),
\@ref(fig:numberSpectraComparison-amidan2014),
\@ref(fig:rtOverTicQuantilesComparison-amidan2014), 
\@ref(fig:rtOverMsQuartersComparison-amidan2014),
\@ref(fig:msSignal10xChangeComparison-amidan2014) 
indicate that the `QuaMeter` and `MsQuality` generally compute
similar values. This is shown by values that show high correlation (points
locate close to the identity line within the scatter plots). We provide in 
Table \@ref(tab:correlation) the Pearson and Spearman correlation coefficients
of the `MsQuality` metrics with their corresponding `QuaMeter` metrics.
Table \@ref(tab:quantiles) is a higher-level analysis regarding the
quantiles of Pearson and Spearman correlation coefficients
between the `MsQuality` and `QuaMeter` metrics. This analysis supports
the observation that `MsQuality` calculates highly similar metric values as
75% of the metrics show Pearson/Spearman coefficients of 0.81/0.87 or higher.

```{r eval = TRUE, echo = FALSE}
.cols <- c("MsQuality", "QuaMeter", "MS level", "Pearson coef.", "Spearman coef.")

df <- data.frame(rbind(
    c("rtDuration", "RT_Duration", 1, 
        round(cor_rtDuration_p$estimate, 3),
        round(cor_rtDuration_s$estimate, 3)), 
    c("numberSpectra", "MS1_Count", 1, 
        round(cor_numberSpectra_msLevel1_p$estimate, 3), 
        round(cor_numberSpectra_msLevel1_s$estimate, 3)),
    c("numberSpectra", "MS2_Count", 2,
        round(cor_numberSpectra_msLevel2_p$estimate, 3), 
        round(cor_numberSpectra_msLevel2_s$estimate, 3)),
    c("rtOverTicQuantiles.25%", "RT_TIC_Q1", 1, 
        round(cor_rt_tic_q1_msLevel1_p$estimate, 3),
        round(cor_rt_tic_q1_msLevel1_s$estimate, 3)),
    c("rtOverTicQuantiles.25%", "RT_TIC_Q1", 2, 
        round(cor_rt_tic_q1_msLevel2_p$estimate, 3),
        round(cor_rt_tic_q1_msLevel2_s$estimate, 3)),
    c("rtOverTicQuantiles.50%", "RT_TIC_Q2", 1, 
        round(cor_rt_tic_q2_msLevel1_p$estimate, 3),
        round(cor_rt_tic_q2_msLevel1_s$estimate, 3)),
    c("rtOverTicQuantiles.50%", "RT_TIC_Q2", 2, 
        round(cor_rt_tic_q2_msLevel2_p$estimate, 3),
        round(cor_rt_tic_q2_msLevel2_s$estimate, 3)),
    c("rtOverTicQuantiles.75%", "RT_TIC_Q3", 1, 
        round(cor_rt_tic_q3_msLevel1_p$estimate, 3),
        round(cor_rt_tic_q3_msLevel1_s$estimate, 3)),
    c("rtOverTicQuantiles.75%", "RT_TIC_Q3", 2, 
        round(cor_rt_tic_q3_msLevel2_p$estimate, 3),
        round(cor_rt_tic_q3_msLevel2_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter1", "RT_MS_Q1", 1, 
        round(cor_rt_msQuarters_q1_msLevel1_p$estimate, 3),
        round(cor_rt_msQuarters_q1_msLevel1_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter1", "RT_MSMS_Q1", 2, 
        round(cor_rt_msQuarters_q1_msLevel2_p$estimate, 3), 
        round(cor_rt_msQuarters_q1_msLevel2_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter2", "RT_MS_Q2", 1, 
        round(cor_rt_msQuarters_q2_msLevel1_p$estimate, 3),
        round(cor_rt_msQuarters_q2_msLevel1_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter2", "RT_MSMS_Q2", 2,
        round(cor_rt_msQuarters_q2_msLevel2_p$estimate, 3),
        round(cor_rt_msQuarters_q2_msLevel2_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter3", "RT_MS_Q3", 1, 
        round(cor_rt_msQuarters_q3_msLevel1_p$estimate, 3),
        round(cor_rt_msQuarters_q3_msLevel1_s$estimate, 3)),
    c("rtOverMsQuarters.Quarter3", "RT_MSMS_Q3", 2,
        round(cor_rt_msQuarters_q3_msLevel2_p$estimate, 3),
        round(cor_rt_msQuarters_q3_msLevel2_s$estimate, 3)),
    c("msSignal10xChange", "IS_1A", 1, 
        round(cor_msSignal10xChange_msLevel1_p$estimate, 3),
        round(cor_msSignal10xChange_msLevel1_s$estimate, 3))
))
colnames(df) <- .cols

knitr::kable(df, caption = "Pearson and Spearman correlation coefficients between \\texttt{MsQuality} and pre-calculated QuaMeter metrics.", format = "latex", label= "correlation")

## calculate some statistics on the correlation values
probs <- c(seq(0, 1, 0.1), 0.25, 0.75) |> 
    sort()
q_pearson <- quantile(as.numeric(df[, "Pearson coef."]), probs = probs) |>
    round(digits = 2)
q_spearman <- quantile(as.numeric(df[, "Spearman coef."]), probs = probs) |>
    round(digits = 2)
df <- data.frame(
    Quantile = names(q_pearson),
    Pearson = q_pearson,
    Spearman = q_spearman
)
rownames(df) <- NULL
colnames(df)[2:3] <- c("Pearson coef.", "Spearman coef.")

knitr::kable(df, caption = "Quantiles for Pearson and Spearman correlation coefficients for \\texttt{MsQuality} and pre-calculated QuaMeter metrics. The correlation analysis showed that 75\\% of the metrics showed Pearson/Spearman correlation coefficients over 0.81/0.87, 50\\% over 0.89/0.91, and 25\\% over 0.96/0.97 between \\texttt{MsQuality} and QuaMeter metrics.", format = "latex", label = "quantiles")

```


\newpage 

## Performance under parallelization

Similar to the above-mentioned analysis using the flow injection
analysis, an important aspect, especially when dealing with large amount of data, 
is scalability and performance when computing the quality metric. 

We measure the time it takes to calculate the quality 
metrics under parallelization of the tasks on 1, 2, 4, 8, and 16 workers 
using the `microbenchmark` package. For computational
reasons we limit the calculation to the first 500 .mzML files.

```{r df_mb_Amidan2014, echo = TRUE, eval = FALSE}
path <- "/scratch/naake/Amidan2014"
fls <- dir(path, full.names = TRUE, recursive = TRUE, pattern = "mzML") |>
    unique()
fls <- fls[1:500]
sps_mb <- sps[sps$dataOrigin %in% fls] 

metrics <- c("rtDuration", "rtOverTicQuantiles", "rtOverMsQuarters", 
    "ticQuartileToQuartileLogRatio", "numberSpectra", 
    "medianPrecursorMz", "rtIqr", "rtIqrRate", "areaUnderTic", 
    "areaUnderTicRtQuantiles", "medianTicRtIqr", "medianTicOfRtRange",
    "mzAcquisitionRange", "rtAcquisitionRange", "precursorIntensityRange", 
    "precursorIntensityQuartiles", "precursorIntensityMean", 
    "precursorIntensitySd", "msSignal10xChange", "ratioCharge1over2", 
    "ratioCharge3over2", "ratioCharge4over2", "meanCharge", 
    "medianCharge")

df_mb <- microbenchmark(
    workers_1 = calculateMetricsFromSpectra(spectra = sps_mb, 
	    metrics = metrics, BPPARAM = MulticoreParam(workers = 1)),
    workers_2 = calculateMetricsFromSpectra(spectra = sps_mb, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 2)),
    workers_4 = calculateMetricsFromSpectra(spectra = sps_mb, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 4)),
    workers_8 = calculateMetricsFromSpectra(spectra = sps_mb, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 8)),
    workers_16 = calculateMetricsFromSpectra(spectra = sps_mb, 
		metrics = metrics, BPPARAM = MulticoreParam(workers = 16)), 
	times = 32L, control = list(warmup = 2), check = "equal"
)
```

```{r df_mb_saveRDS_Amidan2014, eval = FALSE, echo = FALSE}
## save the data.frame
saveRDS(df_mb, file = "../Amidan2014/Amidan2014_df_mb.RDS")
```

```{r df_mb_readRDS_Amidan2014, eval = TRUE, echo = FALSE}
df_mb <- readRDS(file = "../Amidan2014/Amidan2014_df_mb.RDS")
```

```{r microbenchmark-amidan2014, echo = FALSE, eval = TRUE, fig.cap = "Execution time for the calculation of quality metrics of the data set of @Amidan2014 under parallelization (1, 2, 4, 8, and 16 workers).", warning=FALSE}
## convert from nano seconds to seconds
df_mb[["time"]] <- df_mb[["time"]] / 10e9
df_mb <- df_mb |>
    as.data.frame() |>
    mutate(expr = case_when(expr == "workers_1" ~ 1,
        expr == "workers_2" ~ 2, expr == "workers_4" ~ 4,
        expr == "workers_8" ~ 8, expr == "workers_16" ~ 16))

## visualize the elapsed time per amount of used workers
gg_mb <- ggplot(df_mb) +
    geom_jitter(aes(x = expr, y = time, col = as.factor(expr)), alpha = 0.2) +
    geom_violin(aes(x = expr, y = time, group = expr, fill = NA), fill = alpha("white", 0)) +
    xlab("number of workers") + ylab("elapsed time (s)") +
    scale_x_continuous(breaks = df_mb[["expr"]]) +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    theme_classic() +
    theme(legend.position = "none")
gg_mb 
```

The `microbenchmark` package was used to accurately measure the performance 
improvements achieved by parallelization (Figure 
\@ref(fig:microbenchmark-amidan2014)). By parallelizing the calculation of 
the quality metrics across multiple workers, it is possible to significantly 
reduce the execution time.


```{r figure_main, echo = FALSE, message = FALSE, warning=FALSE, results = FALSE, fig.show = "hide"}
## combine several plots into one figure for the main part of the manuscript

## gg_tic_msLevel1: metric for classification into good and poor quality 
## measurements
## gg_tic50_msLevel1: metric for classification into good and poor quality 
## measurements
## gg_rt_tic_msLevel1: comparison between QuaMeter/MsQuality metrics
## gg_mb: benchmark of computation using microbenchmark package
fig_a <- gg_tic_msLevel1 <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
      selected_metric = "areaUnderTic", ylim = c(0, 8e12), 
      plot_margin_1 = unit(c(-15, 0, -15, 0), "cm"), 
      plot_margin_2 = unit(c(-15, 0, -15, 0), "cm"), aspect.ratio = 3.5) |>
    patchworkGrob() |>
    gridExtra::grid.arrange( 
        left = text_grob("area under TIC (A.U.)", size = 10, rot = 90))#, 
        #size = 10, vjust = -0.4)
fig_b <- plot_beeswarm(metrics = metrics_sps_msLevel1, 
    selected_metric = "areaUnderTicRtQuantiles.50%", ylim = c(0, 3e12),
    plot_margin_1 = unit(c(-15, 0, -15, 0), "cm"), 
    plot_margin_2 = unit(c(-15, 0, -15, 0), "cm"), aspect.ratio = 3.5) |>
    patchworkGrob() |>
    gridExtra::grid.arrange( 
        left = text_grob("area under TIC RT quantile (50%)   ", 
            size = 10, rot = 90, just = c("center", "center")))#, 
        #size = 10, vjust = -0.4)
    
    
tmp <- metrics_sps_msLevel1 |>
    filter(!is.na(Curated_Quality)) |>
    mutate(RT_TIC_Q2 = RT_TIC_Q1 + RT_TIC_Q2) |>
    mutate(RT_TIC_Q3 = RT_TIC_Q2 + RT_TIC_Q3) |>
    select(rowname, `rtOverTicQuantiles.25%`, `rtOverTicQuantiles.50%`, `rtOverTicQuantiles.75%`, 
        RT_TIC_Q1, RT_TIC_Q2, RT_TIC_Q3)
tmp <- rbind(
    tmp |>
        select(rowname, `rtOverTicQuantiles.25%`, RT_TIC_Q1) |>
        rename(x = `rtOverTicQuantiles.25%`, y = RT_TIC_Q1) |>
        mutate(variable = "25% quantile"),
    tmp |>
        select(rowname, `rtOverTicQuantiles.50%`, RT_TIC_Q2) |>
        rename(x = `rtOverTicQuantiles.50%`, y = RT_TIC_Q2) |>
        mutate(variable = "50% quantile"),
    tmp |>
        select(rowname, `rtOverTicQuantiles.75%`, RT_TIC_Q3) |>
        rename(x = `rtOverTicQuantiles.75%`, y = RT_TIC_Q3) |>
        mutate(variable = "75% quantile"))
fig_c <- ggplot(tmp, aes(x = x, y = y)) +
        facet_wrap(~ variable) +
        geom_bin_2d(bins = 20) +
        scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), 
            labels = c("0", "0.25", "0.5", "0.75", "1")) +
        xlab("RT over TIC quantiles (MsQuality)") + 
        ylab("RT over TIC quantiles (QuaMeter)") +
        coord_fixed(ratio = 1) +
        theme_classic() +
        theme(legend.position = "none", 
            plot.margin = unit(c(0.2, 0.2, 0.2, 0.5), "cm"))

fig_d <- gg_mb +
    coord_fixed(ratio = 0.042) +
    theme(axis.title = element_text(size = 10), 
        plot.margin = unit(c(-15, 0, -15, 0), "cm"))

pdf("../main/figure-main.pdf")
p1 <- ggarrange(fig_a, fig_b, fig_d, ncol = 3, nrow = 1, labels = c("a", "b", "d"),
    align = "h")
p2 <- ggarrange(fig_c, ncol = 1, nrow = 1, labels = "c")
p1 / p2
dev.off()
```

\newpage

# Session info \label{sec:Session Info}

Information on the attached packages.

```{r sessionInfo, echo = FALSE}
## remove BLAS, LAPACK, and not attached packages
ind_BLAS <- which(names(sessionInfo()) == "BLAS")
ind_LAPACK <- which(names(sessionInfo()) == "LAPACK")
ind_notattached <- which(names(sessionInfo()) == "loadedOnly")
inds_remove <- -c(ind_BLAS, ind_LAPACK, ind_notattached)
utils:::print.sessionInfo(sessionInfo()[inds_remove])
```

\newpage

# References

